\documentclass[11pt,oneside]{book}

\usepackage[letterpaper,margin=1.05in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs,longtable,array}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{float}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{hyperref}

\usetikzlibrary{arrows.meta,positioning,shapes.geometric}
\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.18}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!45!black,
  urlcolor=blue!45!black,
  citecolor=blue!45!black
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}
\setlist[itemize]{leftmargin=1.5em}
\setlist[enumerate]{leftmargin=1.8em}
\sloppy

\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{remark}{Remark}[chapter]
\newtheorem{example}{Example}[chapter]

\newenvironment{proofsteps}
{\begin{enumerate}[label=\textbf{Line \arabic*.}, itemsep=0.35em]}
{\end{enumerate}}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\dd}{\,\mathrm{d}}
\newcommand{\clip}{\operatorname{clip}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\argmin}{\operatorname*{arg\,min}}

\title{\textbf{Monte Carlo Proxy Pricing}\\
\large A textbook-style methodology for European, American, Asian, barrier, cliquet, basket Asian, SLV cliquet, basket cliquet, autocallable, and random-payoff proxies}
\author{Proxy Pricing Research Project}
\date{\today}

\begin{document}
\frontmatter
\maketitle
\tableofcontents

\chapter*{Reader's Guide}
This document is written as a second-year undergraduate text. It assumes calculus, linear algebra, basic probability, and a willingness to read a few stochastic-process formulas. The goal is not to hide the implementation behind jargon. The goal is to explain the finance problem, motivate the proxy solution, show the diagnostic graphs, and then derive the mathematics carefully enough that another researcher can improve the method.

The first six body chapters are intentionally practical:
\begin{itemize}
  \item Motivation: why repeated pricing creates a computational problem.
  \item Solution: how the proxy workflow solves the problem.
  \item Graphs: how to read the value and error diagnostics.
  \item Examples: how the state choice changes by product.
  \item Feature engineering: how to turn financial state into useful proxy inputs.
  \item Method selection: why each proxy method exists and when it should win.
\end{itemize}

The later chapters explain the finance, the mathematics, and the implementation details. The appendices give line-by-line proofs for the formulas used in the code. Every nontrivial implementation trick is classified as one of:
\begin{itemize}
  \item an exact pricing identity,
  \item an unbiased Monte Carlo variance-reduction identity,
  \item a numerical approximation whose error must be validated.
\end{itemize}

\chapter*{Notation and Conventions}
\addcontentsline{toc}{chapter}{Notation and Conventions}
The same notation is used throughout the document.
\begin{longtable}{@{}p{0.22\linewidth}p{0.72\linewidth}@{}}
\toprule
Symbol & Meaning\\
\midrule
\(S_t\) & Spot price at time \(t\). For baskets, \(S_t^{(j)}\) is asset \(j\).\\
\(r,q,\sigma\) & Risk-free rate, dividend yield, and volatility in GBM examples.\\
\(\tau\) & Remaining time to maturity, usually \(T-t\).\\
\(H\) & Final payoff or discounted cash-flow functional before discounting.\\
\(V(t,x)\) & Model value at time \(t\) when the pricing state is \(x\).\\
\(X_t\) & State vector used by the proxy.\\
\(Z\) & Standard-normal simulation shocks after inverse-CDF mapping from Sobol points.\\
\(N\) & Number of Monte Carlo paths or number of fixings, depending on context. The surrounding formula identifies which one.\\
\(A_i,R_i,a_i\) & Running sums: Asian running sum, basket running sum, and cliquet accrued return.\\
\(\ell,u\) & Local floor and cap for a cliquet coupon.\\
\(G_L,G_U\) & Global floor and cap for a cliquet payoff.\\
\(\eta\) & Relative-error denominator floor.\\
\bottomrule
\end{longtable}

All prices in the experiments are model prices. If the simulator is GBM, SLV, or a tree, the proxy is judged against that model. A proxy can be accurate relative to the model and the model can still be a poor market model. The two questions are deliberately separated.

\mainmatter

\chapter{Motivation}

\section{The practical problem}
Modern risk systems often need many derivative prices, not one derivative price. A trader may ask for one price at one market state. A risk engine may need prices across thousands of simulated future states, stress scenarios, portfolio dates, and model configurations.

If one Monte Carlo price costs seconds, then a large exposure run can become too slow. The direct approach is
\[
  \text{market state} \quad \longrightarrow \quad \text{full pricing model} \quad \longrightarrow \quad \text{price}.
\]
The proxy approach inserts an offline learning stage:
\[
  \text{market state} \quad \longrightarrow \quad \text{fast proxy} \quad \longrightarrow \quad \text{approximate model price}.
\]
The aim is not to replace financial modeling. The aim is to preserve the chosen model's price function while making repeated evaluation cheap.

\section{Why this is not ordinary curve fitting}
Derivative prices have financial structure. A good proxy should respect as much of that structure as possible.
\begin{itemize}
  \item A call price should not become negative.
  \item A vanilla call is increasing in spot.
  \item An American value is at least intrinsic value.
  \item A barrier price can change rapidly near the barrier.
  \item A cliquet value has exact floor and cap regions.
  \item A basket product may depend on level, dispersion, correlation, and order statistics.
\end{itemize}
So the project does not ask, ``Which generic machine-learning method fits best?'' The better question is:
\[
  \text{Which financial state, target transform, variance reduction, and fitted proxy work together?}
\]

\section{Success criteria}
The experiments judge a proxy by several criteria, because one number can hide too much.
\begin{itemize}
  \item \emph{Accuracy}: small signed, absolute, and relative errors against an independent benchmark.
  \item \emph{Tail control}: acceptable error in deep in-the-money and out-of-the-money regions.
  \item \emph{Stability}: no artificial oscillation near kinks, barriers, or exercise boundaries.
  \item \emph{Speed}: fast evaluation after training.
  \item \emph{Generality}: the method should extend from European options to path-dependent and higher-dimensional products.
\end{itemize}
The most important accuracy number in this project is often worst-case relative error with a denominator floor. The floor prevents a nearly zero benchmark from producing a meaningless huge percentage.

\chapter{Solution Overview}

\section{The core idea}
For each product, write the expensive model price as a conditional expectation:
\[
  V(t,x)=\E_\Q[Y\mid X_t=x],
\]
where \(Y\) is the discounted future payoff and \(X_t\) is the state. Training means choosing states \(x_i\), estimating labels \(V(t,x_i)\), and fitting a fast approximation \(\widehat V(t,x)\).

The current generic methodology is:
\begin{enumerate}
  \item Use finance to choose a pricing-sufficient state.
  \item Transform the state into numerically stable coordinates.
  \item Generate Monte Carlo labels with Sobol paths, antithetics, control variates, and likelihood-ratio importance sampling when useful.
  \item Use exact regions whenever the payoff logic gives them.
  \item Fit a shape-preserving one-dimensional proxy or a sparse higher-dimensional proxy.
  \item Validate against an independent benchmark.
\end{enumerate}

\section{Workflow graph}
\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.9cm and 0.65cm,
  box/.style={draw, rounded corners=2pt, align=center, minimum width=2.45cm, minimum height=0.78cm, font=\small},
  wide/.style={draw, rounded corners=2pt, align=center, minimum width=3.1cm, minimum height=0.78cm, font=\small},
  arr/.style={-{Latex[length=2mm]}, thick}
]
\node[box] (contract) {Contract\\payoff};
\node[box, right=of contract] (state) {Pricing\\state};
\node[box, right=of state] (labels) {MC labels\\with variance reduction};
\node[box, right=of labels] (fit) {Fit proxy\\target};
\node[box, right=of fit] (validate) {Independent\\benchmark};
\node[wide, below=of labels] (exact) {Exact regions\\bounds and parity};
\node[wide, below=of fit] (report) {Value, signed error,\\relative error, timing};
\draw[arr] (contract) -- (state);
\draw[arr] (state) -- (labels);
\draw[arr] (labels) -- (fit);
\draw[arr] (fit) -- (validate);
\draw[arr] (exact) -- (labels);
\draw[arr] (exact) -- (fit);
\draw[arr] (validate) |- (report);
\draw[arr] (fit) -- (report);
\end{tikzpicture}
\caption{Proxy construction workflow. The fitted method is only one part of the solution; state choice, variance reduction, exact payoff regions, and independent validation are equally important.}
\end{figure}

\section{Default method choices}
For one-feature problems, the current default is SciPy PCHIP on a financially meaningful coordinate. This is used because it is local, shape-preserving, and stable near kinks. For higher-dimensional problems, the default moves toward sparse Chebyshev regression, PCA summaries, moment baselines, and residual calibration. For hard basket cliquet cases, a cached Sobol likelihood-ratio safety proxy is allowed when the fitted surface alone does not meet the error target.

\chapter{Diagnostic Graphs}

\section{Why graphs come before tables}
Tables are useful for comparing many cases, but graphs reveal the failure mode. A method with a good average error can still fail in a wing, near a barrier, or at an exercise kink. For that reason the project uses a three-panel diagnostic:
\begin{itemize}
  \item benchmark value and proxy value;
  \item signed price error, proxy minus benchmark;
  \item signed relative error with a denominator floor.
\end{itemize}
The sign matters. A positive signed error means the proxy overprices relative to the benchmark. A negative signed error means it underprices.

\section{Representative validation dashboard}
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{groupplot}[
  group style={group size=1 by 3, vertical sep=1.25cm},
  width=0.88\linewidth,
  height=0.20\textheight,
  xmin=-3.0, xmax=3.0,
  grid=both,
  major grid style={gray!25},
  minor grid style={gray!12},
  tick label style={font=\small},
  label style={font=\small},
  legend style={font=\small, draw=none, at={(0.02,0.98)}, anchor=north west}
]
\nextgroupplot[
  ylabel={value},
  title={Benchmark value and proxy value}
]
\addplot[very thick, blue!65!black, mark=*] coordinates {
  (-3.0,0.001) (-2.4,0.006) (-1.8,0.040) (-1.2,0.180) (-0.6,0.700) (0.0,1.800)
  (0.6,4.000) (1.2,7.200) (1.8,11.300) (2.4,16.000) (3.0,21.500)
};
\addlegendentry{benchmark}
\addplot[very thick, red!70!black, mark=square*] coordinates {
  (-3.0,0.0012) (-2.4,0.0055) (-1.8,0.038) (-1.2,0.185) (-0.6,0.690) (0.0,1.820)
  (0.6,3.960) (1.2,7.250) (1.8,11.280) (2.4,16.080) (3.0,21.450)
};
\addlegendentry{proxy}

\nextgroupplot[
  ylabel={signed price error},
  title={Proxy minus benchmark},
  scaled y ticks=false,
  yticklabel style={/pgf/number format/fixed, /pgf/number format/precision=2}
]
\addplot[very thick, black, mark=*] coordinates {
  (-3.0,0.0002) (-2.4,-0.0005) (-1.8,-0.0020) (-1.2,0.0050) (-0.6,-0.0100) (0.0,0.0200)
  (0.6,-0.0400) (1.2,0.0500) (1.8,-0.0200) (2.4,0.0800) (3.0,-0.0500)
};
\addplot[dashed, gray] coordinates {(-3,0) (3,0)};

\nextgroupplot[
  ylabel={signed relative error (\%)},
  xlabel={state coordinate, for example \(d_1\) or adjusted moneyness},
  title={Relative error with denominator floor}
]
\addplot[very thick, teal!65!black, mark=triangle*] coordinates {
  (-3.0,0.08) (-2.4,-0.20) (-1.8,-0.80) (-1.2,2.00) (-0.6,-1.43) (0.0,1.11)
  (0.6,-1.00) (1.2,0.69) (1.8,-0.18) (2.4,0.50) (3.0,-0.23)
};
\addplot[dashed, gray] coordinates {(-3,0) (3,0)};
\addplot[dotted, red!60!black] coordinates {(-3,5) (3,5)};
\addplot[dotted, red!60!black] coordinates {(-3,-5) (3,-5)};
\end{groupplot}
\end{tikzpicture}
\caption{Representative diagnostic shape. The coordinates are illustrative, not a fresh benchmark run. The purpose is to show the graph format used throughout the project: values on top, signed price error in the middle, and signed relative error with a practical floor at the bottom.}
\end{figure}

\section{How to interpret the panels}
The top panel answers the first financial question: does the proxy have the same shape as the model value? The middle panel answers the pricing question: where does the proxy overprice or underprice in currency units? The bottom panel answers the risk-control question: is the percentage error acceptable after avoiding meaningless division by tiny values?

For a one-dimensional product, a good proxy usually has a smooth signed-error curve around zero. A sudden spike suggests a missing knot, a bad training label, a barrier/exercise transition, or an extrapolation problem. For a higher-dimensional product, the same idea is applied to slices, stress paths, and validation grids.

\chapter{Guided Examples}

\section{A European call proxy}
\begin{example}[One state variable is enough]
Consider a one-year European call with \(S=100\), \(K=100\), \(r=5\%\), \(q=2\%\), and \(\sigma=20\%\). Under GBM, once today's spot is known, the future distribution of \(S_T\) does not depend on the earlier path. Thus the pricing state is just \(X_t=S_t\).
\end{example}

The proxy should not fit raw spot blindly. A better coordinate is
\[
  d_1
  =
  \frac{\log(S/K)+(r-q+\tfrac12\sigma^2)\tau}{\sigma\sqrt{\tau}}.
\]
This coordinate is financially meaningful because Black-Scholes delta is \(\Phi(d_1)\) for a non-dividend-adjusted call. Spacing points in \(d_1\) gives more balanced coverage of option-delta regions than spacing points uniformly in spot.

\section{An Asian option running-sum example}
\begin{example}[Why the running sum matters]
Suppose an arithmetic Asian call has 12 monthly fixings and strike \(K=100\). Four fixings have already occurred, with running sum
\[
  A=98+101+103+99=401.
\]
Today is a fixing date and today's spot is \(S=105\). There are \(m=7\) future fixings after today. The final average is
\[
  \bar S_T
  =
  \frac{401+105+\sum_{j=1}^{7}S_{t_j}}{12}.
\]
\end{example}

The payoff \((\bar S_T-100)_+\) is equivalent to a scaled option on the future average growth factors. The adjusted strike is
\[
  K_{\rm adj}
  =
  \frac{12\cdot100-401-105}{7}
  =
  99.1429.
\]
The relevant moneyness is therefore \(K_{\rm adj}/S=99.1429/105=0.9442\). This example explains why the proxy uses a coordinate involving both \(S\) and \(A\), not spot alone.

\section{A barrier example}
\begin{example}[Same spot, different barrier state]
Consider a down-and-out call with lower barrier \(L=80\). Two paths can both have today's spot \(S_t=100\). If the first path never touched 80, the option is alive. If the second path touched 80 earlier, the option has already knocked out and is worth zero. The same spot therefore has two different values.
\end{example}

The state must include an alive indicator \(I_t\). A spot-only proxy would average alive and dead cases together, which is financially wrong. This is an example where adding one discrete state variable fixes a serious missing-information problem.

\section{An American put example}
\begin{example}[Exercise versus continuation]
Suppose an American put has strike \(K=100\) and today's spot is \(S=70\). Immediate exercise pays \(K-S=30\). If the discounted expected continuation value is only 28, the value is 30. If the continuation value is 33, the value is 33.
\end{example}

The recursion is
\[
  V(t,S)=\max\{K-S,\;C(t,S)\},
\]
where \(C(t,S)\) is continuation value. A proxy should fit continuation smoothly but impose the max exactly. If it does not impose the max, it can produce values below intrinsic value, which is an arbitrage-style numerical error.

\section{A cliquet exact-tail example}
\begin{example}[No regression needed in a global cap tail]
Suppose a cliquet has global cap \(G_U=20\%\), local floor \(\ell=-2\%\), local cap \(u=4\%\), current accrued return \(a=18\%\), and five coupons remaining. Even if every future coupon is as low as possible, the final accumulated return is at least
\[
  18\%+5(-2\%)=8\%.
\]
So the global cap is not guaranteed. But if \(a=24\%\), then even before future coupons the final clipped payoff is already capped at \(20\%\), so the value is exactly the discounted capped payoff.
\end{example}

Exact regions like this are important. They reduce MC noise and stop the fitted proxy from learning a flat floor or cap region from noisy labels.

\section{A basket Asian example}
\begin{example}[Same basket level, different composition]
Consider a two-asset equal-weight basket. State A has spots \((100,100)\). State B has spots \((130,70)\). Both have basket level 100. If the assets have different volatilities or dividends, or if the payoff depends on nonlinear future averages, the two states need not have the same value.
\end{example}

This explains why a basket proxy often needs composition features, not only basket level. PCA features are a compact way to describe composition and dispersion when the number of underlyings is large.

\section{A basket cliquet order-statistic example}
\begin{example}[Worst-of identity can switch]
Suppose a three-asset worst-of cliquet coupon uses
\[
  \min(R^{(1)},R^{(2)},R^{(3)}).
\]
At one state, asset 1 may be most likely to be worst. At a nearby state, asset 2 may become the likely worst asset. The coupon surface can bend sharply near the switching region.
\end{example}

This is why basket cliquet was harder than European or Asian options. The proxy must handle not only level and volatility but also cross-sectional ranking behavior.

\chapter{Feature Engineering for Proxy Pricing}

\section{Why feature engineering matters}
For proxy pricing, feature engineering is often more important than the choice
between PCHIP, Chebyshev, Akima, or a neural network. A fitting method can only
learn a function of the inputs it receives. If the input omits a state variable
that changes the future payoff distribution, then the approximation error is
structural. More paths and a more powerful regressor cannot fully repair it.

The guiding principle is:
\[
  \text{first make the state financially sufficient, then make it numerically easy.}
\]
A financially sufficient state contains the information needed to price future
cash flows. A numerically easy feature vector rescales and reshapes that state
so that the fitted proxy sees a smoother function.

\section{State variables versus features}
It is useful to separate two concepts:
\begin{itemize}
  \item the \emph{state} is the information needed for pricing;
  \item the \emph{feature vector} is the numerical representation given to the
        fitting method.
\end{itemize}
For a European option under GBM, the state can be \(S_t\), but the feature can
be \(d_1\). For an Asian option, the state can be \((S_t,A_t)\), but the
feature can be adjusted moneyness \(K_{\rm adj}/S_t\). For a basket Asian, the
state contains all spots and the running basket sum, but the feature vector may
contain basket level, running-sum cushion, PCA scores, and a moment baseline.

\section{Feature engineering workflow}
The recommended workflow is:
\begin{enumerate}
  \item Write the payoff and observation schedule.
  \item Identify the Markov or pricing-sufficient state.
  \item Add discrete status variables such as alive, called, knocked out, or
        exercise date.
  \item Find exact regions and bounds before fitting.
  \item Build dimensionless coordinates such as moneyness, log-moneyness,
        \(d_1\), accrued cushion, or normalized variance.
  \item Add baselines, moments, and residual targets when the raw value is too
        curved.
  \item Add cross-sectional features for baskets: level, dispersion, PCA, best,
        worst, and spread indicators.
  \item Validate feature sufficiency by plotting residuals against omitted
        candidate variables.
\end{enumerate}

\section{Common feature types}
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{0.26\linewidth}>{\raggedright\arraybackslash}p{0.31\linewidth}>{\raggedright\arraybackslash}p{0.35\linewidth}@{}}
\toprule
Feature type & Examples & Rationale\\
\midrule
Level & \(S/K\), basket level, forward moneyness & Captures the main price direction.\\
Log level & \(\log(S/K)\), log spot vector & GBM evolves additively in log space.\\
Greek-inspired & \(d_1\), delta bucket, vega scale & Places grid resolution where option sensitivity changes.\\
Path state & running sum, accrued coupon, realized average & Restores sufficiency for path-dependent payoffs.\\
Status flag & alive, called, knocked out, exercise date & Separates states with different payoff rules.\\
Bound/cushion & distance to floor, cap, barrier, strike & Makes exact regions and transition regions visible.\\
Moment baseline & conditional mean, variance, lognormal baseline & Removes the dominant smooth component before fitting.\\
Cross-sectional & basket level, dispersion, best, worst, spread & Captures basket composition and ranking effects.\\
Dimension reduction & PCA scores, common factor, residual factors & Avoids full tensor grids in high dimension.\\
Model state & variance \(v_t\), local-vol bucket, rate state & Includes variables that affect future dynamics.\\
\bottomrule
\end{longtable}

\section{One-feature examples}
\subsection{European option}
Raw spot is a valid state, but \(d_1\) is usually a better feature:
\[
  d_1=
  \frac{\log(S/K)+(r-q+\tfrac12\sigma^2)\tau}{\sigma\sqrt{\tau}}.
\]
The reason is not cosmetic. The Black-Scholes call delta is tied to
\(\Phi(d_1)\), so equal spacing in \(d_1\) gives more balanced coverage across
delta regions than equal spacing in spot.

\subsection{Asian option}
For an arithmetic Asian call with running sum \(A\), current fixing \(S\), and
\(m\) future fixings, the adjusted strike is
\[
  K_{\rm adj}=\frac{NK-A-S}{m}.
\]
The important feature is not \(S\) alone. It is the adjusted moneyness
\[
  \frac{K_{\rm adj}}{S}.
\]
This feature combines today's spot and the running sum into the object that
the remaining future average must beat.

\subsection{Barrier option}
For a barrier option, distance to the barrier is more informative than raw
spot:
\[
  b_{\rm dist}=\log(S/L)
\]
for a lower barrier \(L\). The state also needs the alive flag. Without the
alive flag, two paths with the same spot can have different values.

\subsection{American and Bermudan options}
For early-exercise products, the feature should expose both moneyness and time
to remaining exercise opportunities. A continuation proxy can use spot or
log-moneyness, but the exercise rule must impose
\[
  V=\max(\text{intrinsic},\text{continuation})
\]
after fitting. A feature that hides the exercise boundary tends to create
oscillatory errors.

\section{Basket and high-dimensional examples}
\subsection{Basket Asian}
A basket Asian needs more than basket level. Two states can have the same
basket value but different composition. Useful features include:
\begin{itemize}
  \item current basket level;
  \item running basket sum or remaining average cushion;
  \item weighted log spots;
  \item cross-sectional dispersion;
  \item PCA scores of centered log spots;
  \item conditional mean and variance of the future basket average;
  \item lognormal moment baseline and fitted residual.
\end{itemize}
The moment baseline handles much of the smooth level effect. PCA and
dispersion features explain composition effects that a scalar basket level
cannot see.

\subsection{Basket cliquet}
Basket cliquets need features that describe ranking and dispersion, not only
level. For a worst-of or best-of coupon, the identity of the best or worst
asset can switch. Useful features include:
\begin{itemize}
  \item accrued coupon and remaining floor/cap cushion;
  \item basket return level;
  \item minimum and maximum recent return indicators;
  \item cross-sectional spread;
  \item PCA scores for log spots and log variances;
  \item variance state for each underlying;
  \item local-volatility or SLV state summaries.
\end{itemize}
If the switching surface is too sharp, a fitted smooth proxy may need a cached
Sobol/LR safety layer in that region.

\section{Feature transforms and target transforms}
Feature engineering is linked to target engineering. Some examples:
\begin{itemize}
  \item Use \(\log(V+\epsilon)\) for positive prices with wide dynamic range.
  \item Use a bounded logit target for known floor/cap products.
  \item Use \(\log((V+\epsilon)/(B+\epsilon))\) when a baseline \(B\) captures
        most of the shape.
  \item Use signed or additive residuals when the baseline can be close to
        zero.
\end{itemize}
The feature and target should be chosen together. A good feature can make the
target nearly linear or nearly flat; a poor feature can make even a flexible
regressor work too hard.

\section{How to detect missing features}
After fitting, do not only look at the overall error table. Plot signed errors
against candidate omitted variables. Warning signs include:
\begin{itemize}
  \item residuals that trend with running sum after using spot only;
  \item residuals that trend with distance to barrier;
  \item basket residuals that trend with dispersion or worst-minus-best spread;
  \item cliquet residuals that trend with accrued floor/cap cushion;
  \item SLV residuals that trend with variance after using spot only.
\end{itemize}
If residuals have structure against an omitted variable, the problem is
probably feature engineering, not interpolation.

\section{Feature selection without becoming ad hoc}
The feature set should be product-aware but not arbitrary. A defensible feature
usually comes from one of four sources:
\begin{enumerate}
  \item state sufficiency: the variable is needed by the payoff or dynamics;
  \item asymptotics: the variable controls a known tail, bound, or exact region;
  \item moment approximation: the variable enters conditional mean, variance, or
        a baseline;
  \item validation residuals: independent errors show a stable pattern against
        the variable.
\end{enumerate}
This keeps the methodology generic. It allows product-specific features, but
requires each feature to have a mathematical, financial, or empirical reason.

\section{Practical feature checklist}
Before training a proxy, check:
\begin{itemize}
  \item Does the state determine future payoff distribution under the model?
  \item Are all path-dependent running quantities included?
  \item Are barrier, call, exercise, and knock-out statuses included?
  \item Are features dimensionless and scaled to comparable ranges?
  \item Are known exact regions removed or encoded?
  \item Is there a baseline that can explain the main level effect?
  \item For baskets, are level, dispersion, and ranking effects represented?
  \item For SLV, are spot and variance both represented?
  \item Do residual plots show structure against omitted variables?
  \item Does the feature design still work across parameter cases, not only one
        hand-picked option?
\end{itemize}

\chapter{Method Motivation and Selection}

\section{How to decide that one method is better}
A proxy method is not better in the abstract. It is better for a particular
error mode, product shape, label quality, and dimension. The project therefore
compares methods using several criteria:
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{0.26\linewidth}>{\raggedright\arraybackslash}p{0.66\linewidth}@{}}
\toprule
Criterion & Meaning\\
\midrule
Worst relative error & Controls the bad validation point and matters for production sign-off.\\
p99 relative error & Focuses on tail quality without letting one pathological point dominate.\\
Signed error smoothness & Reveals missing knots, Monte Carlo noise, extrapolation artifacts, and local bias.\\
Shape safety & Avoids negative prices, artificial extrema, and exercise violations.\\
Training cost & Time to generate labels and fit the proxy.\\
Pricing speed & Time to evaluate the proxy after training.\\
Generality & Whether the same idea transfers to the next product family.\\
\bottomrule
\end{longtable}

For PFE and exposure work, shape safety can matter as much as a small average
accuracy improvement. A method that reduces average error but creates a fake
local bump can still be dangerous because exposure quantiles depend on tails
and local maxima.

\section{Decision map}
The current practical map is:
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{0.34\linewidth}>{\raggedright\arraybackslash}p{0.27\linewidth}>{\raggedright\arraybackslash}p{0.31\linewidth}@{}}
\toprule
Situation & First method to try & Reason\\
\midrule
One feature, clean labels, kinks possible & PCHIP & Local, shape-preserving, no artificial extrema.\\
One feature, many slope changes & Akima or MAKIMA & Local cubic slopes react smoothly to uneven local behavior.\\
One feature, very smooth global surface & Chebyshev or Bernstein ridge & Regularized global basis denoises MC labels.\\
One feature, noisy labels & Smoothing spline or P-spline & Trades exact interpolation for lower variance.\\
Early-exercise continuation & PCHIP continuation plus exact max & Avoids oscillatory exercise boundary errors.\\
Known floor or cap & Bounded target plus PCHIP or Chebyshev & Enforces financial bounds by construction.\\
Smooth multi-feature proxy & Sparse Chebyshev ridge & Fast, stable, interpretable, convex fit.\\
Correlated basket proxy & Baseline plus PCA sparse correction & Captures level and composition without a full tensor grid.\\
Order-statistic basket cliquet & Cached Sobol/LR safety proxy & Switching surfaces are too hard for one smooth fitted surface.\\
\bottomrule
\end{longtable}

\section{Linear interpolation}
Linear interpolation is the simplest local baseline. It connects neighboring
labels by straight lines.

It can be better when labels are noisy, when no artificial curvature is
acceptable, or when an audit needs a transparent benchmark. It can be worse
because it is only first-order accurate on smooth curves and can require a
dense grid to match cubic accuracy.

Its main role is diagnostic. If a complicated method performs much worse than
linear interpolation, the complicated method is probably unstable for that
product.

\section{PCHIP}
PCHIP is a local piecewise cubic Hermite interpolant with slope limiting. Its
motivation is financial shape safety: option prices often have monotone
regions, flat tails, exercise kinks, and barrier transitions.

PCHIP can be better because it is local, fast, shape-preserving on monotone
data, and resistant to artificial extrema. In the expanded one-feature study it
had zero local overshoot. It can be worse because it interpolates MC labels
exactly. If labels are visibly noisy, PCHIP preserves that roughness instead of
denoising it.

PCHIP remains the generic one-feature default because it is the best compromise
between accuracy, locality, shape safety, speed, and explainability.

\section{Akima and MAKIMA}
Akima methods also build local cubic interpolants, but their slopes are
weighted from nearby secants rather than limited primarily for monotonicity.
The motivation is to handle local slope changes smoothly.

They can be better when the target has many local slope changes but does not
require strict monotonicity. This is why Akima was strong in the Asian
adjusted-moneyness and random-payoff tests. They can be worse because they are
less strictly shape-preserving than PCHIP and can introduce small overshoot.

Akima is the first local challenger when PCHIP looks too conservative or too
angular.

\section{Natural cubic interpolation}
Natural cubic interpolation passes through all labels and gives a twice
continuously differentiable curve with zero second derivative at the endpoints.

It can be better when the true value surface is smooth and labels are clean.
It was fractionally best on the product-balanced p99 metric in the expanded
one-feature study. It can be worse because it is globally coupled and can
overshoot between labels. That is a serious drawback near barriers, exercise
boundaries, and payoff tails.

The conclusion is that natural cubic interpolation is a credible accuracy
challenger, but not the default for risk use because its small accuracy edge
did not compensate for overshoot risk.

\section{B-spline regression}
B-spline regression represents a curve using local polynomial basis functions.
It is the stable version of regression splines and truncated-power splines.

It can be better when MC labels are noisy and exact interpolation is too
literal. It can place flexibility near important regions through knot choice.
It can be worse because knot placement and regularization become design
choices. If knots miss a barrier or exercise feature, the method can blur the
feature.

B-spline regression is a good denoising candidate, not the simplest production
default.

\section{Smoothing splines and P-splines}
A smoothing spline balances label fit and roughness:
\[
  \sum_i (y_i-f(x_i))^2+\lambda\int (f''(x))^2\,\dd x.
\]
A P-spline uses a B-spline basis with a penalty on coefficient differences.
Both methods are motivated by MC noise reduction.

They can be better when label variance is the main problem. They can be worse
when the smoothing penalty blurs a true kink, barrier transition, or exercise
boundary. Generic smoothing parameters can also underfit steep log-price tails.

The practical use is: try them after improving labels with Sobol, antithetics,
control variates, and importance sampling. They should not be used to hide a
bad simulator.

\section{Adaptive P-splines and local regression}
Adaptive P-splines allow different smoothness in different regions. LOESS fits
local low-degree polynomials around each prediction point. Both are motivated
by varying local behavior.

They can be better near curves with smooth tails and sharp transition regions.
They can be worse because they introduce more hyperparameters and are harder to
audit. LOESS can also be slow at pricing time unless carefully cached.

These methods are useful research tools when ordinary PCHIP or fixed-smoothing
splines leave visible structured errors.

\section{Gaussian-process smoothers}
Gaussian-process smoothers treat the unknown price function as a random smooth
function with a covariance kernel.

They can be better because they provide uncertainty estimates and can guide
adaptive state sampling. They can be worse because training scales poorly with
many states, kernel choices matter, and pricing is slower than interpolation or
linear regression.

Their best role in this project is experimental design: use uncertainty to
choose where to simulate more labels.

\section{Chebyshev ridge}
Chebyshev polynomials are well conditioned on \([-1,1]\), and ridge
regularization stabilizes noisy labels. This makes Chebyshev ridge attractive
for smooth value surfaces.

It can be better because it is fast, convex, stable after scaling, and able to
denoise through regularization. It can be worse near kinks, barriers, and
switching surfaces because it is a global basis. High degree can also overfit
MC noise.

For smooth European-style surfaces and low-dimensional smooth corrections,
Chebyshev ridge is often the strongest global method.

\section{Piecewise Chebyshev}
Piecewise Chebyshev tries to keep spectral accuracy while reducing the global
oscillation problem.

It can be better when different regions of the state space have different
smoothness. It can be worse because the breakpoints, overlaps, and continuity
rules become new modeling choices.

It is a promising bridge between local PCHIP and global Chebyshev, especially
for products with smooth pieces separated by transition regions.

\section{Bernstein and Bezier methods}
Bernstein polynomials have useful shape properties on bounded intervals.
Bezier curves are an equivalent geometric representation of polynomial
segments.

They can be better when positivity, monotonicity, or convexity constraints are
imposed through coefficients. They can be worse when used unconstrained,
because they are still global fits. A piecewise Bezier curve using PCHIP slopes
is not a new method; it is PCHIP written in Bezier form.

Their best future use is constrained regression when the finance gives a known
shape.

\section{Floater-Hormann rational interpolation}
Floater-Hormann interpolation is a pole-free barycentric rational method. It is
motivated by the fact that rational functions can approximate some smooth
curves better than polynomials.

It can be better on difficult smooth approximation problems. It can be worse
for option proxies because it does not automatically respect price bounds or
shape constraints. In the expanded study it produced unacceptable excursions
on some barrier log-value tests.

It is rejected as a generic default.

\section{Sparse Chebyshev for multiple features}
In more than one feature, the full tensor polynomial basis grows too quickly.
Sparse Chebyshev ridge keeps low-order and payoff-relevant interactions.

It can be better because the fit is convex, fast at pricing time,
interpretable, and compatible with analytic derivatives. It can be worse
because the basis terms must be selected carefully. Missing an important
interaction creates bias that more paths will not remove.

The rule is: use sparse Chebyshev only after scaling features and thinking
carefully about which interactions the payoff can create.

\section{Baseline plus residual}
A baseline plus residual proxy uses finance to explain the main price shape and
fits only the correction:
\[
  V(x)=B(x)R(x)
  \qquad\text{or}\qquad
  V(x)=B(x)+\varepsilon_{\rm res}(x).
\]

It can be better because the residual is flatter and lower dimensional than
the raw value. It can be worse if the baseline is poor or nearly zero in an
important region. The ratio target must therefore include a floor.

This is one of the most important ideas for higher-dimensional products:
feature design and baseline design usually matter more than the name of the
regression method.

\section{PCA features}
PCA is motivated by basket structure. A basket can have many spots, but much
of the variation is common level plus a few dispersion directions.

PCA can be better because it reduces dimension while preserving dominant
cross-sectional variation. It can be worse because it is linear and
unsupervised. For order-statistic payoffs, the most important direction may be
an extreme or ranking direction, not the largest-variance direction.

PCA should therefore be used together with payoff-aware features such as basket
level, accrued return, volatility state, and floor/cap cushion.

\section{Neural networks}
Neural networks are flexible high-dimensional function approximators. They are
motivated by products whose interactions are too complex for a small sparse
polynomial basis.

They can be better in genuinely high dimension, especially with large training
sets. They can be worse because they are harder to audit, need architecture and
training choices, and do not automatically respect financial constraints.

They are a future benchmark for high-dimensional exotics, not the current
default.

\section{Cached Sobol/LR safety proxy}
Some products are too hard for one fitted smooth surface. Basket cliquet
order-statistic coupons can switch which asset is best or worst. A cached
Sobol likelihood-ratio safety proxy prices difficult states by reusing a strong
simulator rather than relying on a weak fitted extrapolation.

It can be better because it is much safer in discontinuous or switching
regions. It can be worse because it is slower than an analytic fitted proxy and
still has MC error.

This is the current safety fallback for basket cliquet hard cases.

\section{GPU acceleration}
GPU acceleration is not a new pricing theory and not a new proxy family. It is
a computational backend choice. The important question is where the parallel
work actually lives.

The strongest GPU candidate is Monte Carlo label generation:
\[
  x_i \longmapsto
  \frac1N\sum_{n=1}^N e^{-r\tau}H(x_i,Z_n).
\]
For each state \(x_i\), many paths can be simulated independently. For each
path, many assets, time steps, payoff variants, and likelihood-ratio weights
can be computed in parallel. This is exactly the kind of work GPUs are good at.

The same argument applies to independent benchmark generation. If a benchmark
requires hundreds of thousands or millions of paths at many validation states,
then the benchmark is often more GPU-friendly than the final fitted proxy.

\begin{longtable}{@{}>{\raggedright\arraybackslash}p{0.30\linewidth}>{\raggedright\arraybackslash}p{0.28\linewidth}>{\raggedright\arraybackslash}p{0.34\linewidth}@{}}
\toprule
Task & GPU usefulness & Rationale\\
\midrule
MC training labels & High & Paths, states, assets, and payoff variants are massively parallel.\\
MC benchmarks & High & Same structure as training labels, often with even more paths.\\
Basket Asian simulation & High & Many assets, correlations, and state points.\\
SLV and basket cliquet simulation & High & Many time steps, variance states, coupons, and LR weights.\\
Barrier simulation & Medium to high & Path monitoring and Brownian-bridge factors are parallel, but branch behavior near barriers needs care.\\
PCHIP, Akima, linear interpolation & Low & Sorting and slope construction are small CPU tasks.\\
Small Chebyshev ridge fits & Low to medium & CPU is usually enough unless the design matrix is large or many fits are batched.\\
Large sparse basis evaluation & Medium & GPU can help build \(A\), \(A^\top A\), and batched predictions.\\
Neural-network training & High & Matrix operations and minibatch training are GPU-native.\\
\bottomrule
\end{longtable}

The current practical recommendation is therefore not ``move everything to the
GPU.'' It is to use the GPU for labels and heavy high-dimensional training,
while keeping small deterministic fitting and reporting on CPU.

There is one important caveat. This project prefers Sobol and other
low-discrepancy designs for MC labels. GPU pseudo-random normals are easy, but
replacing Sobol with pseudo-random paths just to use the GPU could increase
variance. A careful GPU plan should test one of the following:
\begin{itemize}
  \item generate Sobol points on CPU and transfer them to GPU in chunks;
  \item use a GPU library with reliable Sobol or quasi-Monte Carlo support;
  \item compare GPU pseudo-random antithetic labels against CPU Sobol labels at
        equal error, not merely equal path count.
\end{itemize}

Likelihood-ratio mixtures also need numerical care on GPU. It is safer to
compute likelihood weights in log space and normalize only when necessary,
because tail shifts can create very small or very large weights.

The proposed architecture is a backend abstraction:
\[
  \operatorname{simulate\_labels}(\text{contract},\text{states},\text{config},\text{backend}).
\]
The backend choices could include CPU Sobol, GPU Sobol, and GPU
pseudo-random antithetic modes. The validation report should compare
wall time, standard error, and final proxy error for each backend.

This is still open for discussion. GPU acceleration is most attractive if the
training or benchmark time is dominated by paths. It is less attractive if the
run is dominated by Python overhead, file output, small spline fits, or model
logic with heavy branching.

\section{Recommended method ladder}
Before inventing a new method, use this sequence:
\begin{enumerate}
  \item Verify that the state is pricing-sufficient.
  \item Add exact payoff regions: maturity, intrinsic value, barrier hit state,
        floor/cap tails, and in/out parity.
  \item Improve labels using Sobol paths, antithetics, control variates, common
        random numbers, and LR mixture importance sampling.
  \item Choose a financially scaled coordinate such as \(d_1\), adjusted
        moneyness, accrued cushion, normalized variance, or PCA score.
  \item Try PCHIP for one feature and sparse Chebyshev for several smooth
        features.
  \item If labels are noisy, test smoothing spline or P-spline.
  \item If local slope changes dominate, test Akima or MAKIMA.
  \item If the target is globally smooth, test Chebyshev or Bernstein ridge.
  \item If the product is high-dimensional, add baseline plus residual and PCA.
  \item If strict max error still fails, use adaptive sampling or a cached
        Sobol/LR safety proxy in the hard region.
  \item If label or benchmark generation dominates runtime, test a GPU MC
        backend while preserving Sobol/LR accuracy controls.
\end{enumerate}

\section{Further improvements}
The most valuable next improvements are:
\begin{itemize}
  \item adaptive state sampling driven by validation error;
  \item state-dependent MC path allocation;
  \item shape-constrained splines or Bernstein fits where monotonicity,
        convexity, or bounds are known;
  \item adaptive sparse Chebyshev basis selection;
  \item product-aware LR mixture components;
  \item GPU backends for MC label generation, benchmarks, and neural-network
        training;
  \item outer-scenario weighted validation for PFE;
  \item neural-network benchmarks for genuinely high-dimensional products.
\end{itemize}

\chapter{Financial Background}

\section{What a derivative price means}
A derivative is a contract whose cash flows are determined by market variables such as spot prices, interest rates, volatilities, realized averages, barrier events, or coupon-reset rules. A European call, for example, pays \((S_T-K)_+\) at maturity. An Asian option pays a function of an average. A cliquet pays a function of capped and floored period returns.

The mathematical payoff is not yet the price. A price is the amount of money at the valuation date that is economically equivalent to the future cash flows under the chosen pricing model. In a complete frictionless model, this equivalence is understood through replication: if a self-financing trading strategy produces the same future cash flow as the derivative, then the derivative must have the same initial cost as that strategy. Otherwise an arbitrage would exist.

For this project, the word \emph{price} always means \emph{model price}. The model may be GBM, SLV, a tree, a finite-difference method, or a Monte Carlo simulator. The proxy is judged by how well it approximates the model price. It does not by itself prove that the model is a good description of the market.

\section{Why risk-neutral pricing appears}
It is tempting to think that a derivative price is the discounted real-world expected payoff,
\[
  e^{-r(T-t)}\E_{\mathbb P}[H\mid\mathcal F_t],
\]
where \(\mathbb P\) is the real-world probability measure. That is usually wrong. Risky cash flows earn risk premia, and the physical drift of the underlying asset is not the right drift for no-arbitrage valuation.

Risk-neutral pricing changes the probability measure from \(\mathbb P\) to \(\Q\) so that discounted tradable asset prices are martingales. In the Black-Scholes model, this means the stock drift becomes \(r-q\), not the investor's expected return \(\mu\). The pricing formula then becomes
\[
  V_t
  =
  B_t\,\E_\Q\!\left[\frac{H}{B_T}\,\middle|\,\mathcal F_t\right].
\]
This is not saying investors are risk-neutral. It is saying that, after risk has been priced through the change of measure, the derivative can be valued as a discounted expectation under \(\Q\).

\begin{remark}[Model choice and market calibration]
In practice, desks calibrate models to liquid market instruments before using them for less liquid instruments. The proxy work in this repository starts after that modeling choice has been made. The central question is: given a pricing model, can we replace repeated expensive valuation with a fast and accurate approximation?
\end{remark}

\section{Why proxies are useful}
Monte Carlo pricing can be expensive because one price may require many simulated paths. A risk report, PFE calculation, XVA run, stress test, or calibration loop may need thousands or millions of prices. A proxy replaces the expensive pricing function
\[
  x \longmapsto V(t,x)
\]
with a cheaper function
\[
  x \longmapsto \widehat V(t,x).
\]
The proxy is trained offline using expensive labels and then used online for fast repeated evaluation.

This separation is the main engineering tradeoff:
\[
  \text{spend time once to train} \quad \Longrightarrow \quad \text{save time many times later}.
\]
The proxy is valuable only if it preserves the financial shape of the price: monotonicity, convexity when relevant, exercise kinks, barrier transitions, floor/cap regions, and tail behavior.

\section{Product families and path dependence}
The products in this project differ mainly in what information from the past is needed to value the future.

\begin{longtable}{@{}p{0.22\linewidth}p{0.34\linewidth}p{0.34\linewidth}@{}}
\toprule
Product & Financial feature & State idea\\
\midrule
European & payoff depends only on terminal spot & current spot is enough under GBM\\
Asian & payoff depends on an average & current spot plus running sum\\
Basket Asian & payoff depends on an average basket & all current spots plus running basket sum\\
American/Bermudan & holder may exercise early & current spot and exercise date\\
Barrier & payoff depends on whether a level was hit & current spot plus alive/hit status\\
Cliquet & payoff depends on capped/floored reset returns & accrued coupon plus market state\\
Autocallable & early redemption at observation dates & spot, date, and alive status\\
\bottomrule
\end{longtable}

This is why proxy design is not only a regression problem. It is first a financial-state problem. If the state omits information that affects future cash flows, then no interpolation method can fully repair the error.

\section{Model risk, numerical risk, and proxy risk}
There are three different errors that should not be confused.
\begin{itemize}
  \item \emph{Model risk}: the chosen dynamics may not match the market.
  \item \emph{Numerical risk}: the benchmark simulator may have time-step, path-count, or discretization error.
  \item \emph{Proxy risk}: the fitted approximation may differ from the benchmark model value.
\end{itemize}
This repository mainly studies proxy risk and numerical risk under specified models. When closed-form benchmarks exist, as for European Black-Scholes options, they are used. When they do not exist, the benchmark is a stronger numerical calculation, usually with more paths, independent randomization, and variance reduction.

\chapter{Mathematical Background}

\section{Conditional expectation as the pricing target}
The central mathematical object is a conditional expectation:
\[
  V(t,x)=\E_\Q[Y\mid X_t=x],
\]
where \(Y\) is the discounted future payoff and \(X_t\) is the state. This formula says: among all scenarios that share the same current state \(x\), average the discounted future payoff under the pricing measure.

For a second-year undergraduate, the useful intuition is that conditional expectation is the best mean-square prediction using the information currently available. If \(g(X_t)\) is any function of the state, then
\[
  \E\!\left[(Y-\E[Y\mid X_t])^2\right]
  \le
  \E\!\left[(Y-g(X_t))^2\right].
\]
So proxy pricing can be viewed as approximating the best possible state-based prediction of the discounted payoff.

\section{Pricing-sufficient states}
\begin{definition}[Pricing-sufficient state]
Let \((\Omega,\mathcal F,(\mathcal F_t)_{t\ge0},\Q)\) be a filtered risk-neutral probability space, and let \(H\) be an integrable payoff. An adapted process \(X_t\in\R^d\) is called pricing-sufficient for \(H\) at time \(t\) if there exists a measurable function \(v\) such that
\[
  B_t\,\E_\Q\!\left[\frac{H}{B_T}\,\middle|\,\mathcal F_t\right]
  =
  v(t,X_t).
\]
\end{definition}

This definition is the professional version of the earlier informal phrase ``a good state is sufficient.'' It says exactly what must be true: after conditioning on \(X_t\), no missing history can change the price. The state may still be too high-dimensional for convenient fitting, but it is at least financially correct.

\begin{remark}[Sufficient does not mean minimal]
A sufficient state may contain redundant variables. For example, including both spot and log spot is unnecessary but still sufficient. A minimal state removes redundancy, but finding the truly minimal state is often less important than finding a stable low-dimensional state that prices accurately.
\end{remark}

\section{Markov models and state evolution}
A process is Markov if the conditional law of the future depends on the past only through the current state. In a one-asset GBM model,
\[
  \frac{\dd S_t}{S_t}=(r-q)\dd t+\sigma\dd W_t,
\]
the future distribution of \(S_T\) depends on the current history only through \(S_t\). Thus \(S_t\) is a pricing-sufficient state for European options.

For path-dependent products, the spot alone is usually not Markov for the payoff. An Asian option also needs the running sum. A barrier option needs to know whether the barrier has already been hit. A cliquet needs the accrued coupon. The mathematical task is to enlarge the state just enough to restore the Markov property for the future cash flows.

\section{Brownian motion, GBM, and lognormality}
GBM is used throughout the simpler examples because it has a closed-form transition:
\[
  S_{t+\Delta t}
  =
  S_t
  \exp\!\left((r-q-\tfrac12\sigma^2)\Delta t+\sigma\sqrt{\Delta t}\,Z\right),
  \qquad Z\sim N(0,1).
\]
This formula explains why log-moneyness and \(d_1\)-type variables are natural coordinates. The randomness enters the log price additively, so a coordinate based on \(\log(S/K)\) is better scaled than raw spot.

For multiple assets, the normal shocks are correlated. If \(\rho\) is the correlation matrix and \(L L^\top=\rho\), then \(LZ\) has correlation \(\rho\) when \(Z\) has independent standard-normal components. This is the linear algebra behind basket simulations.

\section{Dynamic programming for exercise and callable products}
Early-exercise and callable products are valued backward in time. At a decision date, the value is the better of stopping and continuing. For an American-style product,
\[
  V_k(x)=\max\{g_k(x), C_k(x)\},
\]
where \(g_k(x)\) is immediate exercise value and
\[
  C_k(x)=e^{-r\Delta t}\E_\Q[V_{k+1}(X_{k+1})\mid X_k=x]
\]
is continuation value.

The proxy problem is therefore not only terminal payoff fitting. It can also be continuation-value fitting. This is why monotone and shape-preserving one-dimensional methods are useful: exercise boundaries create kinks, and a global polynomial can oscillate near those kinks.

\section{Approximation viewpoint}
After MC labels are generated, fitting is an approximation problem. There are three common viewpoints.
\begin{itemize}
  \item \emph{Interpolation}: pass exactly through relatively clean labels. PCHIP is in this category.
  \item \emph{Regression}: fit noisy labels with a lower-dimensional function class. Ridge Chebyshev regression is in this category.
  \item \emph{Baseline plus correction}: use finance to explain most of the price, then fit the smaller residual.
\end{itemize}
The best method depends on dimension and noise. In one feature, shape-preserving interpolation is hard to beat when labels are accurate. In many features, sparse regression, PCA summaries, and baseline corrections become more important because grids become too large.

\section{Curse of dimensionality}
If one dimension uses \(m\) grid points, a full tensor grid in \(d\) dimensions uses \(m^d\) points. With \(m=20\) and \(d=10\), this is \(20^{10}\), which is impossible for routine training. This is the curse of dimensionality.

The project responds to this in several ways:
\begin{itemize}
  \item reduce the state using finance, such as adjusted moneyness for Asian options;
  \item use PCA to summarize correlated cross-sectional variation;
  \item use sparse Chebyshev bases rather than full tensor bases;
  \item use baselines so that the fitted residual is simpler than the original price;
  \item use cached MC safety pricing for hard basket cliquet cases where a fitted proxy alone is not reliable enough.
\end{itemize}

\chapter{The Pricing Problem}

\section{Risk-neutral value}
Let \(H\) be the payoff of a derivative at maturity or final settlement time \(T\). Under the risk-neutral measure \(\Q\), the time-\(t\) value is
\[
  V_t = B_t\,\E_\Q\!\left[\frac{H}{B_T}\,\middle|\,\mathcal F_t\right],
\]
where \(B_t=\exp(\int_0^t r_s\dd s)\) is the money-market account. If the short rate is constant,
\[
  V_t = e^{-r(T-t)}\E_\Q[H\mid\mathcal F_t].
\]

In the code, the expensive object is the conditional expectation. The proxy is a fast approximation of the map
\[
  x \longmapsto V(t,x),
\]
where \(x\) is the state available at the valuation date.

\section{What makes a good proxy state}
A state is good when it is sufficient, low-dimensional, and scaled sensibly.

\begin{definition}[Sufficient state]
A vector \(X_t\) is sufficient for pricing if, after conditioning on \(X_t\), the distribution of all future cash flows no longer depends on earlier path history.
\end{definition}

Examples:
\begin{itemize}
  \item European option: \(X_t=S_t\).
  \item Arithmetic Asian option: \(X_t=(S_t,A_t)\), where \(A_t\) is the running sum of past fixings.
  \item American put: \(X_t=S_t\), but the time index matters because of remaining exercise dates.
  \item Barrier option: \(X_t=(S_t,I_t)\), where \(I_t\) says whether the barrier has already been hit.
  \item Single-name SLV cliquet: \(X_t=(a_t,S_t,v_t)\), accrued coupon, spot, and variance.
  \item Basket SLV cliquet: \(X_t=(a_t,S_t^{(1:3)},v_t^{(1:3)})\), plus contract variant.
\end{itemize}

\section{Error decomposition}
The observed proxy error is not one number with one cause. A useful decomposition is
\[
  \widehat V_{\mathrm{proxy}} - V_{\mathrm{true}}
  =
  \underbrace{\widehat V_{\mathrm{proxy}} - V_{\mathrm{MC-label}}}_{\text{fitting error}}
  +
  \underbrace{V_{\mathrm{MC-label}}-V_{\mathrm{model}}}_{\text{training MC noise}}
  +
  \underbrace{V_{\mathrm{model}}-V_{\mathrm{true}}}_{\text{model/discretization error}}.
\]
The experiments in this repository mostly study the first two terms under fixed model assumptions.

\chapter{Monte Carlo Label Construction}

\section{Plain estimator}
For a state \(x\), define the discounted path value
\[
  Y(x,Z)=e^{-r\tau}H(x,Z),
\]
where \(Z\) represents all future random shocks. The target label is
\[
  V(x)=\E[Y(x,Z)].
\]
With \(N\) simulated paths,
\[
  \widehat V_N(x)=\frac1N\sum_{n=1}^N Y(x,Z_n).
\]
The estimator is unbiased when the paths have the correct law. Its standard error is estimated by
\[
  \widehat{\se}(\widehat V_N)
  =
  \sqrt{\frac{1}{N(N-1)}\sum_{n=1}^N\bigl(Y_n-\overline Y\bigr)^2}.
\]

\section{Sobol and antithetics}
Sobol points reduce integration error by filling the unit cube more evenly than independent random points. The implementation rounds the number of paths to a power of two because Sobol nets are naturally balanced in blocks of \(2^m\).

Antithetic pairing uses the fact that, for a standard normal vector \(Z\),
\[
  Z \stackrel{d}= -Z.
\]
The paired payoff
\[
  \frac{f(Z)+f(-Z)}{2}
\]
has the same expectation as \(f(Z)\). It often has lower variance for monotone or nearly monotone payoffs.

\section{Default simulation policy}
The default policy used by the project is:
\begin{enumerate}
  \item Use Sobol low-discrepancy points whenever the path dimension is known in advance.
  \item Use antithetic pairing when the payoff simulator accepts both \(Z\) and \(-Z\).
  \item Use a likelihood-ratio shift or shift mixture for tail-sensitive labels whenever a useful direction is available.
  \item Use the unshifted component as part of the mixture unless a closed-form variance calculation says a pure shift is safe.
  \item Use independent randomization or independent Sobol scrambles for benchmarks.
\end{enumerate}
The phrase ``whenever possible'' matters. For example, if a payoff is synthetic and the useful tail direction is unknown, the safe choice is a mixture containing negative, zero, and positive shifts. If a benchmark uses a finite-difference or tree method, no MC importance sampling is needed because there is no path simulation in that benchmark.

\section{From Sobol uniforms to GBM paths}
For a \(d\)-asset GBM model over one step \(\Delta t\), start with a Sobol vector
\[
  U=(U_1,\ldots,U_d)\in(0,1)^d.
\]
Map it to independent normals by
\[
  Z_i=\Phi^{-1}(U_i).
\]
If \(L\) is a Cholesky factor of the correlation matrix \(\rho\), so that \(LL^\top=\rho\), set
\[
  \widetilde Z=LZ.
\]
Then the simulated next spot is
\[
  S_{t+\Delta t}^{(j)}
  =
  S_t^{(j)}
  \exp\!\left((r-q_j-\tfrac12\sigma_j^2)\Delta t
       +\sigma_j\sqrt{\Delta t}\,\widetilde Z_j\right).
\]
For a path with many dates, repeat this step date by date. The Brownian increments are independent across time before correlation is applied across assets.

\section{Likelihood-ratio importance sampling}
State enrichment and importance sampling are different.

State enrichment changes the distribution of training states \(x_i\). It does not change the path law for any individual conditional expectation.

Importance sampling changes the path density from \(p\) to \(q\) and multiplies by a likelihood ratio:
\[
  \E_p[h(Z)]
  =
  \E_q\!\left[h(Z)\frac{p(Z)}{q(Z)}\right].
\]
For a normal mean shift, \(q(z)=\phi(z-\theta)\), so
\[
  \frac{p(z)}{q(z)}
  =
  \exp\!\left(-\theta z+\frac12\theta^2\right).
\]
This identity is exact; its usefulness depends on variance.

\section{Mixture importance sampling}
A single shift can be too opinionated. The more generic tail-biased method is a mixture:
\[
  q(z)=\sum_{m=1}^M \pi_m \phi_d(z-\theta_m),
  \qquad
  \pi_m>0,\quad \sum_{m=1}^M\pi_m=1.
\]
Here \(\phi_d\) is the \(d\)-dimensional standard normal density and \(\theta_m\) is a vector shift. The likelihood ratio is
\[
  L(z)=\frac{\phi_d(z)}{\sum_{m=1}^M\pi_m\phi_d(z-\theta_m)}.
\]
The weighted estimator is
\[
  \widehat V_{\rm IS}
  =
  \frac1N\sum_{n=1}^N h(Z_n)L(Z_n),
  \qquad Z_n\sim q.
\]
For option training, mixture shifts are useful because one model can include lower-tail, upper-tail, and central scenarios without needing a different formula for every product. The cost is that likelihood weights can become variable; therefore the code tracks benchmark error and avoids using an aggressive shift as the only source of paths.

\section{Control variates}
If \(C\) has known mean \(m_C\), then
\[
  Y_\beta = Y-\beta(C-m_C)
\]
has the same expectation as \(Y\). The variance-minimizing coefficient is
\[
  \beta^\star=\frac{\Cov(Y,C)}{\Var(C)}.
\]
Asian options use a geometric Asian control when the arithmetic payoff and geometric payoff are strongly correlated.

\chapter{Proxy Fitting Principles}

\section{Targets}
Raw prices can be hard to fit because they range from nearly zero to almost linear. The code uses transformations such as:
\[
  \log(V+\varepsilon),\qquad
  \log\frac{V+\varepsilon}{B+\varepsilon},\qquad
  \operatorname{logit}\frac{V-L}{U-L}.
\]
Here \(B\) is a baseline, and \(L,U\) are known lower and upper bounds.

\section{Exact regions}
Never ask a regressor to learn something known analytically.

Examples:
\begin{itemize}
  \item At maturity, use the payoff exactly.
  \item Deep Asian call in-the-money region: the positive part disappears and the value is linear.
  \item Cliquet global floor/cap tails: the final clipped payoff is already known.
  \item American/Bermudan exercise: impose \(V=\max(\text{intrinsic},\text{continuation})\).
\end{itemize}

\section{Relative error floor}
The reported relative error is
\[
  e_{\mathrm{rel}}
  =
  \frac{|\widehat V-V|}{\max(|V|,\eta)}.
\]
This is ordinary relative error when \(|V|\ge \eta\), and it avoids meaningless exploding percentages when \(V\) is tiny.

\chapter{One-Dimensional Shape-Preserving Proxies}

\section{Why PCHIP is the default for 1D}
For a single feature, PCHIP is a strong operational default:
\begin{itemize}
  \item it interpolates labels exactly;
  \item it is local, so one noisy point does not create global oscillation;
  \item it preserves monotone shape when the data are monotone;
  \item it behaves well near exercise kinks, barrier transitions, and low-value tails.
\end{itemize}

\section{Cubic Hermite form}
On \([x_i,x_{i+1}]\), let \(h_i=x_{i+1}-x_i\), \(t=(x-x_i)/h_i\). A cubic Hermite segment is
\[
\begin{aligned}
H_i(t)
&=y_i(2t^3-3t^2+1)
  +h_i d_i(t^3-2t^2+t)\\
&\quad
  +y_{i+1}(-2t^3+3t^2)
  +h_i d_{i+1}(t^3-t^2).
\end{aligned}
\]
PCHIP chooses slopes \(d_i\) from neighboring secants in a way that prevents artificial extrema.

\section{PCHIP slope construction}
Let
\[
  \delta_i=\frac{y_{i+1}-y_i}{x_{i+1}-x_i}
\]
be the secant slope. At an interior knot \(x_i\), PCHIP applies this rule.
\begin{itemize}
  \item If \(\delta_{i-1}\delta_i\le0\), set \(d_i=0\). This prevents a spurious turn at a sign change.
  \item If \(\delta_{i-1}\) and \(\delta_i\) have the same sign, set
  \[
    d_i=
    \frac{w_1+w_2}{w_1/\delta_{i-1}+w_2/\delta_i},
    \qquad
    w_1=2h_i+h_{i-1},\quad
    w_2=h_i+2h_{i-1}.
  \]
\end{itemize}
The second formula is a weighted harmonic mean. A harmonic mean is pulled toward the smaller magnitude, which stops the interpolant from using a large slope to overshoot a flat part of the curve.

Endpoint slopes use one-sided information and are then clipped to respect the first interior secant. This is why SciPy's \texttt{PchipInterpolator} is preferred in the main code: it gives a mature, tested implementation of the Fritsch-Butland endpoint and interior rules.

\section{Why exact interpolation is acceptable}
PCHIP interpolates labels exactly, so it is not a smoother. That is acceptable only when MC noise is already controlled by Sobol paths, antithetics, control variates, and enough paths. If labels are visibly noisy, a smoothing spline or local regression can have lower average error. The project therefore uses PCHIP as the default one-feature method but still treats label construction as the first-order accuracy problem.

\section{Akima, splines, and Bernstein}
Akima interpolation is also local and robust to outliers. Natural cubic splines are smoother but can overshoot. Bernstein polynomials are shape-friendly on bounded intervals and were competitive for some barrier variants. For one-dimensional production defaults, the current rule is:
\[
  \text{PCHIP first, compare Akima/Bernstein when the validation surface says so.}
\]

\chapter{Sparse Chebyshev and Higher Dimension}

\section{Chebyshev scaling}
Chebyshev polynomials satisfy
\[
  T_n(\cos\theta)=\cos(n\theta),
\]
so \(|T_n(x)|\le 1\) for \(x\in[-1,1]\). This is why features are scaled before fitting:
\[
  z_j = 2\frac{x_j-\ell_j}{u_j-\ell_j}-1.
\]

\section{Ridge regression}
Given a design matrix \(A\), labels \(y\), and penalty matrix \(\Gamma\), ridge regression solves
\[
  \min_\beta \|A\beta-y\|_2^2+\lambda\|\Gamma\beta\|_2^2.
\]
The normal equation is
\[
  (A^\top A+\lambda\Gamma^\top\Gamma)\beta=A^\top y.
\]

\section{Sparse multi-index basis}
For a \(D\)-dimensional state \(z=(z_1,\ldots,z_D)\in[-1,1]^D\), a tensor Chebyshev basis term is
\[
  \Phi_\alpha(z)=\prod_{j=1}^D T_{\alpha_j}(z_j),
  \qquad
  \alpha=(\alpha_1,\ldots,\alpha_D).
\]
A full tensor basis with \(\alpha_j\le p\) is usually too large. A total-degree basis keeps only
\[
  |\alpha|_1=\alpha_1+\cdots+\alpha_D\le p.
\]
The implementation also allows payoff-aware pruning. For example, a basket Asian correction keeps level, dispersion, running-sum, and a few PCA interaction terms before it considers high-order cross-products. This is not a pricing theorem; it is a bias-variance decision that must be checked by validation.

\section{Baseline plus residual}
High-dimensional products are easier to fit when the proxy is not asked to learn the whole price from scratch. The generic decomposition is
\[
  V(x)=B(x)\,R(x),
\]
or, with an additive residual,
\[
  V(x)=B(x)+\varepsilon_{\rm res}(x).
\]
Here \(B(x)\) is a cheap baseline: Black-Scholes, moment-matched lognormal, intrinsic value plus discounted floor, or a low-dimensional approximation. The fitted target can then be
\[
  y(x)=\log\frac{V(x)+\epsilon}{B(x)+\epsilon}.
\]
When \(B\) captures most of the shape, the residual is flatter and lower dimensional. This is the main reason the basket Asian and basket cliquet proxies improved after adding moment, PCA, and grouped-label features.

\section{Why sparsity is necessary}
A full tensor degree-\(p\) basis in \(D\) variables contains \((p+1)^D\) terms. With \(D=7\) and \(p=5\), that is \(279{,}936\) terms before fitting. Basket cliquet therefore needs sparse terms, payoff-aware features, or a slower safety method.

\chapter{Instrument Recipes}

\section{Recipe template}
Each product section answers the same six questions:
\begin{enumerate}
  \item What is the payoff?
  \item What state is sufficient for future cash flows?
  \item Which coordinate is used by the proxy?
  \item How are MC labels generated?
  \item What benchmark is used?
  \item Which parts are exact identities and which parts are fitted approximations?
\end{enumerate}
This template is important because a proxy can look accurate for the wrong reason. For example, reusing training paths in the benchmark can hide error, and using a coordinate that is not sufficient can look fine on a small grid but fail when parameters change.

\section{European option}
State: \(S\). Coordinate:
\[
  d_1=\frac{\log(S/K)+(r-q+\tfrac12\sigma^2)\tau}{\sigma\sqrt{\tau}}.
\]
Training labels use shifted Sobol terminal GBM paths. Validation uses Black-Scholes:
\[
  C=S e^{-q\tau}\Phi(d_1)-K e^{-r\tau}\Phi(d_2),\qquad d_2=d_1-\sigma\sqrt{\tau}.
\]
Default proxy: log price fitted by PCHIP in \(d_1\).

Exact parts: the terminal GBM formula and the Black-Scholes benchmark. Fitted part: the interpolation from finitely many noisy labels. Main failure mode: tiny option values can make raw percentage error unstable, so validation reports both signed price error and relative error with a floor.

\section{Arithmetic Asian option}
State: \(S\) and running sum \(A\). With \(m\) future fixings and total \(N\),
\[
  K_{\mathrm{adj}}=\frac{NK-A-S}{m}.
\]
The call payoff can be rewritten as
\[
  \left(\frac{A+S+\sum_{j=1}^m S G_j}{N}-K\right)_+
  =
  S\frac{m}{N}\left(\frac1m\sum_{j=1}^mG_j-\frac{K_{\mathrm{adj}}}{S}\right)_+.
\]
Thus the tested monthly call can be represented by an adjusted-moneyness coordinate. Training uses Sobol antithetics, likelihood shifts where useful, and a geometric Asian control variate.

Exact parts: the payoff rewrite and deterministic in-the-money or out-of-the-money regions. Unbiased parts: Sobol randomized labels, antithetics, and likelihood weights. Approximate parts: using a one-dimensional adjusted-moneyness fit for the tested family and using MC for the arithmetic-average benchmark.

\section{Ten-asset basket Asian}
State: ten spots plus the running basket sum. The proxy starts from a moment-matched lognormal baseline for the final arithmetic basket average, then fits a sparse Chebyshev/PCA correction and a PCHIP residual calibration. PCA summarizes cross-sectional composition without throwing away the level and running-sum information.

Exact parts: first and second basket moments under GBM. Approximate parts: the lognormal moment match and the sparse residual proxy. The moment match is not expected to be exact because an arithmetic basket average is a sum of correlated lognormal variables, and that sum is not lognormal.

\section{American and Bermudan options}
At each exercise date \(k\),
\[
  V_k(S)=\max\!\left(g_k(S),\,e^{-r\Delta t}\E[V_{k+1}(S_{k+1})\mid S_k=S]\right).
\]
The continuation value is estimated by one-step Sobol transitions and fitted with PCHIP. The max with intrinsic value is imposed exactly.

Exact parts: dynamic-programming recursion and the exercise max. Approximate parts: finite exercise grid for Bermudan/American discretization, MC continuation labels, and interpolation of continuation. Benchmarks use a tree or projected finite-difference method to avoid testing the proxy against itself.

\section{Barrier option}
For a zero-rebate barrier,
\[
  V_{\mathrm{in}}+V_{\mathrm{out}}=V_{\mathrm{vanilla}}.
\]
Discrete monitoring checks simulated observation dates. Continuous monitoring uses Brownian-bridge survival between observations. For a lower log barrier \(b\), endpoints \(x,y>b\), and variance \(\sigma^2\Delta t\),
\[
  P(\text{survive}\mid x,y)=1-\exp\!\left(-\frac{2(x-b)(y-b)}{\sigma^2\Delta t}\right).
\]

Exact parts: discrete barrier path payoff and in/out parity. Conditional part: the Brownian-bridge survival probability is exact for continuous monitoring under log-GBM between two fixed endpoints. Approximate parts: MC integration and interpolation across spot.

\section{Cliquet and SLV cliquet}
Cliquet payoffs are bounded by local and global caps/floors. If accrued value \(a\) and \(m\) remaining coupons imply the final total is certainly below the global floor or above the global cap, the value is exact.

For single-name SLV cliquet, the state is \((a,S,v)\). The proxy uses exact tails, bounded-logit targets, and local/sparse Chebyshev features. For three-asset basket cliquet, the fitted proxy is not universally enough for strict max-error control; hard order-statistic cases use a cached Sobol/LR safety proxy.

Exact parts: local coupon clipping, global floor/cap clipping, and tail bounds. Approximate parts: SLV time stepping, sparse fitted surfaces, and any cached MC safety proxy. The safety proxy is slower than a closed-form fitted proxy, but it is still much faster than regenerating a full benchmark for every repeated query.

\section{Autocallable and random payoff}
Autocallable valuation is naturally date-indexed. The state is spot at an observation date; the cash-flow rule determines whether the note redeems early or continues. Random payoff tests are synthetic one-dimensional payoffs used to stress interpolation under different shapes.

Exact parts: the call/no-call cash-flow rule at an observation date. Approximate parts: continuation labels and interpolation. The synthetic random payoff tests are not financial products; they are diagnostic shapes used to catch overshoot, poor extrapolation, and kink handling.

\chapter{Contract Definitions and Payoff Algebra}

\section{European call and put}
For a call and put with strike \(K\),
\[
  H_{\rm call}=(S_T-K)_+,\qquad
  H_{\rm put}=(K-S_T)_+.
\]
The proxy problem is one-dimensional because, under constant-parameter GBM, all future uncertainty depends on the current state only through the current spot \(S_t\). In practice the coordinate \(d_1\), rather than raw spot, is preferred because equal spacing in \(d_1\) allocates nodes by option delta rather than by currency units of spot.

\section{Arithmetic Asian call}
Let \(0=t_0<t_1<\cdots<t_{N-1}=T\) be fixing dates. At valuation date \(t_i\), let
\[
  A_i=\sum_{j=0}^{i-1} S_{t_j}
\]
be the running sum before today's fixing. If today's spot is \(S\), then the final arithmetic average is
\[
  \bar S_T=\frac{A_i+S+\sum_{k=i+1}^{N-1}S_{t_k}}{N}.
\]
The payoff is
\[
  H=(\bar S_T-K)_+.
\]
The state is two-dimensional before reduction: \(X_i=(S,A_i)\). The adjusted-strike algebra in Appendix C explains why the tested call can be fit on a one-dimensional adjusted-moneyness coordinate.

\section{Basket Asian call}
For \(d\) assets with basket weights \(w_j\),
\[
  B_{t_k}=\sum_{j=1}^d w_j S_{t_k}^{(j)},\qquad
  \bar B_T=\frac{R_i+B_{t_i}+\sum_{k=i+1}^{N-1}B_{t_k}}{N},
\]
where \(R_i\) is the running sum of past basket fixings. The payoff is
\[
  H=(\bar B_T-K)_+.
\]
This is no longer reducible to one scalar in general because the future basket distribution depends on the full cross-section of spots and correlations. The code therefore uses moment baselines and PCA features.

\section{Barrier call}
For a knock-out call with lower barrier \(L\) and upper barrier \(U\), the alive indicator is
\[
  I_T=
  \mathbf 1\{L<S_{t_k}<U \text{ for all monitored } t_k\}.
\]
The payoff is
\[
  H_{\rm out}=I_T(S_T-K)_+.
\]
For a knock-in, \(H_{\rm in}=(1-I_T)(S_T-K)_+\), so \(H_{\rm in}+H_{\rm out}=H_{\rm vanilla}\) path by path.

\section{Cliquet payoff}
For one underlying, the local return over period \(k\) is
\[
  R_k=\frac{S_{t_k}}{S_{t_{k-1}}}-1.
\]
The locally clipped coupon is
\[
  C_k=\clip(R_k,\ell,u).
\]
If \(a_i=\sum_{k=1}^{i}C_k\) is the accrued clipped return at reset \(i\), the final payoff is
\[
  H=N_0\,\clip\!\left(a_i+\sum_{k=i+1}^{M}C_k,\;G_L,\;G_U\right),
\]
where \(N_0\) is notional and \(G_L,G_U\) are global floor and cap.

\section{Basket cliquet coupon variants}
Let \(R_k^{(j)}\) be the \(j\)-th asset return in period \(k\). The repository tests several coupon maps. Representative examples are
\[
\begin{aligned}
C_k^{\rm basket}
  &= \clip\!\left(\sum_j w_j R_k^{(j)},\ell,u\right),\\
C_k^{\rm worst}
  &= \clip\!\left(\min_j R_k^{(j)},\ell,u\right),\\
C_k^{\rm best}
  &= \clip\!\left(\max_j R_k^{(j)},\ell,u\right),\\
C_k^{\rm spread}
  &= \clip\!\left(\sum_j w_jR_k^{(j)},\ell,u\right)
     -\lambda_s\clip\!\left(\max_jR_k^{(j)}-\min_jR_k^{(j)},s_L,s_U\right)
     +b\,\mathbf 1_{\{\sum_jw_jR_k^{(j)}\ge0\}}.
\end{aligned}
\]
Order-statistic coupons are hard because small changes in cross-sectional dispersion can change which asset is best or worst.

\section{Autocallable payoff recursion}
At observation date \(k\), if the note has not called and \(S_{t_k}\ge B_{\rm call}K\), the contract redeems. A schematic recursion is
\[
  V_k(S)=
  \begin{cases}
  N_0(1+c_k), & S\ge B_{\rm call}K,\\
  e^{-r\Delta t}\E[V_{k+1}(S_{k+1})\mid S_k=S], & S<B_{\rm call}K.
  \end{cases}
\]
At maturity the redemption rule may include coupon, capital protection, or downside participation.

\chapter{Implementation Algorithms}

\section{Generic proxy construction algorithm}
For every instrument, the implementation follows this structure.
\begin{enumerate}
  \item Define the contract payoff and observation schedule.
  \item Prove or choose a Markov state \(X_t\).
  \item Transform the state to dimensionless coordinates: moneyness, \(d_1\), normalized accrued return, normalized variance, PCA scores, or expected-total cushion.
  \item Generate a training state grid. Include broad coverage and extra nodes near exercise, barrier, floor/cap, and low-value transition regions.
  \item Generate Monte Carlo labels with Sobol, antithetics, control variates, and likelihood-ratio shifts where they lower variance.
  \item Fit a proxy target such as \(\log(V+\varepsilon)\), a bounded logit, or a baseline residual.
  \item Validate on independent states and independent random streams.
  \item Report values, signed errors, absolute errors, relative errors with a floor, and benchmark standard errors when available.
\end{enumerate}

\section{European algorithm}
\begin{enumerate}
  \item Fix \(\tau=T-t\).
  \item Choose training nodes \(S_i=S(d_{1,i})\) with \(d_{1,i}\) covering low and high deltas.
  \item For each node, estimate \(V_i\) by shifted Sobol terminal paths:
  \[
    S_T^{(i,n)}
    =
    S_i\exp\!\left((r-q-\tfrac12\sigma^2)\tau+\sigma\sqrt\tau\,Z_{i,n}^{\rm shifted}\right).
  \]
  \item Fit \(y_i=\log(V_i+\varepsilon)\) with PCHIP in \(d_1\).
  \item Predict by \(\widehat V(S)=\exp(\widehat y(d_1(S)))-\varepsilon\), clipped at zero.
  \item Benchmark by Black-Scholes.
\end{enumerate}

\section{Asian algorithm}
\begin{enumerate}
  \item For each fixing date, build states \((S,A)\) in adjusted moneyness and running-average directions.
  \item Compute \(K_{\rm adj}=(NK-A-S)/m\).
  \item Use exact linear/zero regions when \(K_{\rm adj}\le0\) or the option is certainly out of the money.
  \item For the active region, simulate future fixings by Sobol antithetic GBM paths.
  \item Estimate the arithmetic payoff and geometric payoff on the same paths.
  \item Use the exact geometric Asian value as a control variate.
  \item Fit normalized value and time value with PCHIP.
  \item Validate with an independent high-path benchmark.
\end{enumerate}

\section{Basket Asian algorithm}
\begin{enumerate}
  \item Build states \(X=(S^{(1)},\ldots,S^{(10)},R)\).
  \item Compute the mean and variance of the final arithmetic basket average.
  \item Price the moment-matched lognormal baseline \(B(X)\).
  \item Compute PCA scores of centered log spots to summarize composition.
  \item Fit a sparse Chebyshev correction to \(\log((V+\varepsilon)/(B+\varepsilon))\).
  \item Bin residuals against expected-average moneyness and fit a PCHIP residual calibration.
  \item Validate against Sobol/LR benchmarks on independent state grids.
\end{enumerate}

\section{American and Bermudan algorithm}
\begin{enumerate}
  \item Set terminal value equal to payoff.
  \item Move backward through exercise dates.
  \item At each date, simulate one-step transitions from training spots.
  \item Evaluate the already-fitted next-date value on simulated next spots.
  \item Average and discount to estimate continuation.
  \item Fit continuation by PCHIP.
  \item Define \(V_k(S)=\max(g_k(S),\widehat C_k(S))\).
  \item Validate American values against finite differences and Bermudan values against a tree.
\end{enumerate}

\section{Barrier algorithm}
\begin{enumerate}
  \item Build an alive-state spot grid inside the barrier domain.
  \item Simulate discrete monitoring paths with Sobol antithetics.
  \item For continuous monitoring, multiply each path by Brownian-bridge survival probabilities on each segment.
  \item Fit the alive knock-out value by the best validated one-dimensional interpolator.
  \item Obtain knock-in values by in/out parity when needed.
  \item Validate by independent high-path simulation.
\end{enumerate}

\section{Cliquet and basket cliquet algorithm}
\begin{enumerate}
  \item Represent the state by accrued return and any market variables needed for the next returns.
  \item Use exact global floor/cap tails before simulation.
  \item Simulate remaining local returns under GBM or SLV.
  \item Reuse each simulated future coupon sum for several accrued values when possible.
  \item Fit a bounded target in logit space.
  \item For high-dimensional order-statistic basket coupons, fall back to cached Sobol/LR safety pricing when fitted surfaces fail max-error targets.
\end{enumerate}

\chapter{Validation and Timing}

\section{Independent benchmark principle}
Training labels and benchmark labels use separate random streams. Benchmarks use more paths, closed forms, finite differences, or tree methods when available.

\section{Current reporting}
The timing report separates:
\begin{itemize}
  \item training sample generation;
  \item proxy fitting after labels are available;
  \item option-family wall time;
  \item average wall time per scenario.
\end{itemize}
The reduced-budget smoke accuracy columns in the timing report are diagnostic only. The product-specific markdown summaries remain the full-budget accuracy source.

\appendix

\chapter{Line-by-Line Proofs: Conditional Pricing and State Choice}

\section{Risk-neutral pricing formula}
\begin{theorem}[Risk-neutral valuation, constant short rate]
If the discounted asset gains are martingales under \(\Q\) and the payoff \(H\) is integrable, then the time-\(t\) model value is
\[
  V_t=e^{-r(T-t)}\E_\Q[H\mid\mathcal F_t].
\]
\end{theorem}
\begin{proofsteps}
\item The money-market account is \(B_t=e^{rt}\).
\item Absence of arbitrage in the risk-neutral model says the discounted value process \(B_t^{-1}V_t\) is a \(\Q\)-martingale.
\item Therefore
\[
  B_t^{-1}V_t
  =
  \E_\Q[B_T^{-1}H\mid\mathcal F_t].
\]
\item Since \(B_t=e^{rt}\) and \(B_T=e^{rT}\),
\[
  e^{-rt}V_t=e^{-rT}\E_\Q[H\mid\mathcal F_t].
\]
\item Multiply both sides by \(e^{rt}\) to obtain the formula.
\end{proofsteps}

\section{When a proxy state is sufficient}
\begin{theorem}[State reduction by conditional distribution]
Suppose \(X_t\) is \(\mathcal F_t\)-measurable and the conditional distribution of future cash flows \(H\) given \(\mathcal F_t\) depends on the past only through \(X_t\). Then there exists a function \(v\) such that
\[
  V_t=v(t,X_t).
\]
\end{theorem}
\begin{proofsteps}
\item Define the price by conditional expectation:
\[
  V_t=e^{-r(T-t)}\E[H\mid\mathcal F_t].
\]
\item By assumption, once \(X_t\) is known, earlier path details do not change the conditional law of \(H\).
\item Therefore \(\E[H\mid\mathcal F_t]\) is a function of \(X_t\).
\item Call this function \(g(t,X_t)\).
\item Set \(v(t,X_t)=e^{-r(T-t)}g(t,X_t)\).
\item Hence \(V_t=v(t,X_t)\).
\end{proofsteps}

\section{Tower property for backward induction}
\begin{theorem}[Tower property]
If \(\mathcal G\subseteq\mathcal F\), then
\[
  \E[\E[Y\mid\mathcal F]\mid\mathcal G]=\E[Y\mid\mathcal G].
\]
\end{theorem}
\begin{proofsteps}
\item The conditional expectation \(\E[Y\mid\mathcal F]\) is the best \(\mathcal F\)-measurable summary of \(Y\) in the sense of matching integrals over \(\mathcal F\)-events.
\item Every \(\mathcal G\)-event is also an \(\mathcal F\)-event because \(\mathcal G\subseteq\mathcal F\).
\item Therefore replacing \(Y\) by \(\E[Y\mid\mathcal F]\) preserves all integrals over \(\mathcal G\)-events.
\item The conditional expectation with respect to \(\mathcal G\) is the unique \(\mathcal G\)-measurable variable with those integrals.
\item Hence the identity follows.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Monte Carlo}

\section{Unbiased sample mean}
\begin{proofsteps}
\item Define \(Y=e^{-r\tau}H\) and \(V=\E_\Q[Y]\).
\item Define \(\widehat V_N=N^{-1}\sum_{n=1}^N Y_n\).
\item Apply linearity:
\[
\E[\widehat V_N]
=\E\!\left[\frac1N\sum_{n=1}^N Y_n\right]
=\frac1N\sum_{n=1}^N\E[Y_n]
=\frac1N\sum_{n=1}^N V
=V.
\]
\item If paths are independent with variance \(\sigma_Y^2\),
\[
\Var(\widehat V_N)
=\Var\!\left(\frac1N\sum_{n=1}^N Y_n\right)
=\frac1{N^2}\sum_{n=1}^N\Var(Y_n)
=\frac{\sigma_Y^2}{N}.
\]
\item Replace \(\sigma_Y^2\) by the sample variance to estimate standard error.
\end{proofsteps}

\section{Likelihood-ratio shift}
\begin{proofsteps}
\item Start with \(V=\int h(z)p(z)\dd z\).
\item Choose a proposal \(q\) with \(q(z)>0\) wherever \(h(z)p(z)\neq 0\).
\item Multiply and divide by \(q\):
\[
V=\int h(z)\frac{p(z)}{q(z)}q(z)\dd z
=\E_q[h(Z)L(Z)],\qquad L(Z)=\frac{p(Z)}{q(Z)}.
\]
\item For \(p(z)=\phi(z)\), \(q(z)=\phi(z-\theta)\),
\[
L(z)
=\frac{\exp(-z^2/2)}{\exp(-(z-\theta)^2/2)}
=\exp\!\left(-\theta z+\frac{\theta^2}{2}\right).
\]
\item For independent shocks, likelihood ratios multiply:
\[
L(z_1,\ldots,z_m)
=\prod_{j=1}^m\exp\!\left(-\theta_j z_j+\frac{\theta_j^2}{2}\right).
\]
\end{proofsteps}

\section{Mixture likelihood ratio}
\begin{proofsteps}
\item Let the original density be \(p(z)\).
\item Choose shifted proposal densities \(q_m(z)\) and probabilities \(\pi_m\) with \(\sum_m\pi_m=1\).
\item Define the mixture density
\[
  q(z)=\sum_{m=1}^M\pi_m q_m(z).
\]
\item Require \(q(z)>0\) wherever \(h(z)p(z)\neq0\).
\item Multiply and divide by \(q\):
\[
  \int h(z)p(z)\dd z
  =
  \int h(z)\frac{p(z)}{q(z)}q(z)\dd z.
\]
\item Therefore, if \(Z\sim q\),
\[
  \E_p[h(Z)]
  =
  \E_q[h(Z)L(Z)],
  \qquad
  L(Z)=\frac{p(Z)}{q(Z)}.
\]
\item To simulate from \(q\), first draw a component \(M\) with probability \(\pi_m\), then draw \(Z\sim q_M\).
\item The likelihood ratio must use the full mixture denominator \(q(z)\), not only the chosen component density \(q_M(z)\), because the sampled value \(z\) could also have been generated by other components.
\end{proofsteps}

\section{Antithetic estimator}
\begin{proofsteps}
\item Standard normal symmetry gives \(Z\stackrel d=-Z\).
\item Define \(A=(f(Z)+f(-Z))/2\).
\item Then
\[
\E[A]=\frac{\E[f(Z)]+\E[f(-Z)]}{2}
=\frac{\E[f(Z)]+\E[f(Z)]}{2}
=\E[f(Z)].
\]
\item The variance is
\[
\Var(A)=\frac12\Var(f(Z))+\frac12\Cov(f(Z),f(-Z)).
\]
\item The pairing helps whenever the covariance term is negative.
\end{proofsteps}

\chapter{Line-by-Line Proofs: GBM and European Options}

\section{GBM solution}
\begin{proofsteps}
\item Start from \(\dd S_t/S_t=(r-q)\dd t+\sigma\dd W_t\).
\item Apply Ito's lemma to \(f(S)=\log S\):
\[
\dd\log S_t
=\frac1{S_t}\dd S_t-\frac12\frac1{S_t^2}(\dd S_t)^2.
\]
\item Since \((\dd S_t)^2=\sigma^2S_t^2\dd t\),
\[
\dd\log S_t=(r-q-\tfrac12\sigma^2)\dd t+\sigma\dd W_t.
\]
\item Integrate over \([t,T]\):
\[
\log(S_T/S_t)=(r-q-\tfrac12\sigma^2)\tau+\sigma(W_T-W_t).
\]
\item Use \(W_T-W_t=\sqrt{\tau}Z\), \(Z\sim N(0,1)\), and exponentiate.
\end{proofsteps}

\section{Black-Scholes call}
\begin{proofsteps}
\item Risk-neutral pricing:
\[
C=e^{-r\tau}\E[(S_T-K)_+].
\]
\item Split the positive part:
\[
\E[(S_T-K)_+]
=\E[S_T\mathbf 1_{\{S_T>K\}}]-K\Q(S_T>K).
\]
\item The event \(S_T>K\) is \(Z>-d_2\), so \(\Q(S_T>K)=\Phi(d_2)\).
\item Completing the square gives
\[
\E[S_T\mathbf 1_{\{S_T>K\}}]
=S e^{(r-q)\tau}\Phi(d_1).
\]
\item Substitute and discount:
\[
C=S e^{-q\tau}\Phi(d_1)-K e^{-r\tau}\Phi(d_2).
\]
\end{proofsteps}

\chapter{Line-by-Line Proofs: Asian State Reduction}

\section{Adjusted moneyness}
\begin{proofsteps}
\item At a valuation date, write future fixings as \(S G_1,\ldots,S G_m\).
\item The arithmetic average is
\[
\bar S=\frac{A+S+S\sum_{j=1}^mG_j}{N}.
\]
\item Define
\[
K_{\rm adj}=\frac{NK-A-S}{m}.
\]
\item Rewrite the call payoff:
\[
(\bar S-K)_+
=\frac1N\left(A+S+S\sum_{j=1}^mG_j-NK\right)_+
=\frac{m}{N}\left(S\frac1m\sum_{j=1}^mG_j-K_{\rm adj}\right)_+.
\]
\item Factor out \(S>0\):
\[
(\bar S-K)_+
=S\frac{m}{N}\left(\frac1m\sum_{j=1}^mG_j-\frac{K_{\rm adj}}{S}\right)_+.
\]
\item Under GBM the growth factors \(G_j\) do not depend on the current spot level \(S\).
\item Therefore \(V(S,A)=S F(K_{\rm adj}/S,t)\), which justifies a one-dimensional adjusted-moneyness proxy for the tested call setup.
\end{proofsteps}

\section{Geometric Asian control variate}
For future fixing times \(t_1,\ldots,t_m\), define the geometric average
\[
  G=\left(\prod_{j=1}^m S_{t_j}\right)^{1/m}.
\]
\begin{proofsteps}
\item Take logs:
\[
  \log G=\frac1m\sum_{j=1}^m \log S_{t_j}.
\]
\item Under GBM, each \(\log S_{t_j}\) is normal.
\item A linear combination of jointly normal random variables is normal.
\item Therefore \(\log G\sim N(\mu_G,\sigma_G^2)\).
\item The mean is
\[
  \mu_G
  =
  \log S_0
  +
  \frac1m\sum_{j=1}^m(r-q-\tfrac12\sigma^2)t_j.
\]
\item The variance is
\[
  \sigma_G^2
  =
  \frac{\sigma^2}{m^2}
  \sum_{i=1}^m\sum_{j=1}^m \min(t_i,t_j),
\]
because \(\Cov(W_{t_i},W_{t_j})=\min(t_i,t_j)\).
\item Since \(G\) is lognormal, the call value \(\E[(G-K)_+]\) has the same algebraic form as Black-Scholes with log-mean \(\mu_G\) and log-variance \(\sigma_G^2\).
\item This exact geometric Asian value can be used as the known mean \(m_C\) in the control-variate formula.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Barriers}

\section{In/out parity}
\begin{proofsteps}
\item Split all paths into two disjoint events: hit and no hit.
\item On hit paths, the knock-in pays the vanilla payoff and the knock-out pays zero.
\item On no-hit paths, the knock-out pays the vanilla payoff and the knock-in pays zero.
\item Therefore, path by path,
\[
H_{\rm in}+H_{\rm out}=H_{\rm vanilla}.
\]
\item Discount and take expectations:
\[
V_{\rm in}+V_{\rm out}=V_{\rm vanilla}.
\]
\end{proofsteps}

\section{Brownian-bridge lower-barrier survival}
\begin{theorem}[Brownian-bridge crossing probability]
Let \(X_s\) be a Brownian bridge from \(x\) to \(y\) over length \(\Delta t\), with diffusion variance \(\sigma^2\). If \(x>b\) and \(y>b\), then
\[
  \Q\!\left(\min_{0\le s\le\Delta t}X_s\le b\,\middle|\,X_0=x,X_{\Delta t}=y\right)
  =
  \exp\!\left(-\frac{2(x-b)(y-b)}{\sigma^2\Delta t}\right).
\]
\end{theorem}
\begin{proofsteps}
\item Condition on log endpoints \(X_0=x\), \(X_{\Delta t}=y\), both above lower barrier \(b\).
\item By the reflection principle, the conditional probability of crossing below \(b\) is
\[
\exp\!\left(-\frac{2(x-b)(y-b)}{\sigma^2\Delta t}\right).
\]
\item Survival is one minus crossing:
\[
P_{\rm surv}=1-\exp\!\left(-\frac{2(x-b)(y-b)}{\sigma^2\Delta t}\right).
\]
\item Conditional survival over many intervals is the product of interval survival probabilities.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Exercise and Cliquet Bounds}

\section{American recursion}
\begin{proofsteps}
\item At final date \(M\), \(V_M(S)=g_M(S)\).
\item Suppose \(V_{k+1}\) is known.
\item Continuing from date \(k\) has value
\[
C_k(S)=e^{-r\Delta t}\E[V_{k+1}(S_{k+1})\mid S_k=S].
\]
\item Stopping immediately has value \(g_k(S)\).
\item The holder chooses the larger:
\[
V_k(S)=\max(g_k(S),C_k(S)).
\]
\item Backward induction proves optimality at every date.
\end{proofsteps}

\section{Cliquet exact tails}
\begin{proofsteps}
\item Let current accrued return be \(a\), with \(m\) future local coupons.
\item Each future coupon lies in \([\ell,u]\).
\item Therefore final accumulated return lies in
\[
a+\sum_{j=1}^m c_j \in [a+m\ell,\;a+mu].
\]
\item If \(a+mu\) is below the global floor, the payoff is exactly the global floor.
\item If \(a+m\ell\) is above the global cap, the payoff is exactly the global cap.
\item These states need no regression and no simulation.
\end{proofsteps}

\chapter{Line-by-Line Proofs: PCHIP}

\section{Unique Hermite cubic}
\begin{proofsteps}
\item A cubic has four coefficients.
\item Endpoint value constraints give two linear equations.
\item Endpoint slope constraints give two more linear equations.
\item The Hermite basis constructs one solution:
\[
H(t)=y_i h_{00}(t)+h_i d_i h_{10}(t)+y_{i+1}h_{01}(t)+h_i d_{i+1}h_{11}(t).
\]
\item If two cubics satisfied all four conditions, their difference would have double roots at both endpoints.
\item A nonzero cubic cannot have four roots counting multiplicity, so the solution is unique.
\end{proofsteps}

\section{Why PCHIP avoids artificial extrema}
\begin{theorem}[Fritsch-Carlson monotonicity criterion]
For a cubic Hermite interpolant on a monotone interval, if the normalized endpoint slopes satisfy the Fritsch-Carlson inequalities, then the derivative does not change sign inside the interval. Hence the cubic segment is monotone on that interval.
\end{theorem}
\begin{proofsteps}
\item Let \(\delta_i=(y_{i+1}-y_i)/h_i\).
\item If neighboring secants change sign, PCHIP sets the node slope to zero.
\item If neighboring secants have the same sign, PCHIP uses the weighted harmonic mean
\[
d_i=\frac{w_1+w_2}{w_1/\delta_{i-1}+w_2/\delta_i}.
\]
\item This slope has the same sign as the neighboring secants and lies between their magnitudes.
\item On a monotone interval, the derivative of the Hermite cubic is a quadratic whose normalized endpoint slopes satisfy the Fritsch-Carlson monotonicity inequalities.
\item Therefore the derivative does not change sign inside the interval.
\item Hence the interpolant cannot create a new local maximum or minimum between two monotone data points.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Sparse Chebyshev Ridge}

\section{Chebyshev boundedness}
\begin{proofsteps}
\item Every \(x\in[-1,1]\) can be written as \(x=\cos\theta\).
\item Define \(T_n(x)=\cos(n\theta)\).
\item Since \(|\cos(n\theta)|\le1\),
\[
|T_n(x)|\le1,\qquad x\in[-1,1].
\]
\item Scaling state variables to \([-1,1]\) keeps basis columns numerically bounded.
\end{proofsteps}

\section{Ridge normal equation}
\begin{proofsteps}
\item Objective:
\[
J(\beta)=\|A\beta-y\|_2^2+\lambda\|\Gamma\beta\|_2^2.
\]
\item Differentiate:
\[
\nabla J(\beta)=2A^\top(A\beta-y)+2\lambda\Gamma^\top\Gamma\beta.
\]
\item Set the gradient to zero:
\[
A^\top A\beta-A^\top y+\lambda\Gamma^\top\Gamma\beta=0.
\]
\item Rearrange:
\[
(A^\top A+\lambda\Gamma^\top\Gamma)\beta=A^\top y.
\]
\item The left matrix is positive semidefinite, and ridge makes unsupported directions stable.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Bounded Targets and Metrics}

\section{Logit bounded target}
\begin{proofsteps}
\item Suppose the price must lie in \([L,U]\).
\item Fit an unconstrained number \(y\) and map it through the logistic function:
\[
u(y)=\frac{1}{1+e^{-y}}.
\]
\item Because \(e^{-y}>0\), \(0<u(y)<1\).
\item Rescale:
\[
\widehat V=L+(U-L)u(y).
\]
\item Therefore \(L<\widehat V<U\).
\end{proofsteps}

\section{Relative-error floor}
\begin{proofsteps}
\item Define
\[
e_{\rm rel}=\frac{|\widehat V-V|}{\max(|V|,\eta)},\qquad \eta>0.
\]
\item If \(|V|\ge\eta\), this is ordinary relative error.
\item If \(|V|<\eta\), then
\[
e_{\rm rel}=\frac{|\widehat V-V|}{\eta}.
\]
\item Thus a nearly zero benchmark cannot create an infinite percentage error.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Basket Moments and PCA}

\section{Basket Asian first moment}
Let the basket fixing at time \(t_k\) be
\[
  B_{t_k}=\sum_{j=1}^d w_j S_{t_k}^{(j)}.
\]
Under correlated risk-neutral GBM,
\[
  \E[S_{t_k}^{(j)}\mid S_0^{(j)}]
  =
  S_0^{(j)}e^{(r-q_j)t_k}.
\]
Therefore
\[
  \E[B_{t_k}]
  =
  \sum_{j=1}^d w_j S_0^{(j)}e^{(r-q_j)t_k}.
\]
For the final arithmetic basket average
\[
  \bar B=\frac{R_i+B_{t_i}+\sum_{k=i+1}^{N-1}B_{t_k}}{N},
\]
linearity gives
\[
  \E[\bar B]
  =
  \frac{R_i+B_{t_i}+\sum_{k=i+1}^{N-1}\E[B_{t_k}]}{N}.
\]

\section{Basket Asian second moment}
For two assets \(j,\ell\) and times \(t_a,t_b\), correlated GBM gives
\[
\E[S_{t_a}^{(j)}S_{t_b}^{(\ell)}]
=
S_0^{(j)}S_0^{(\ell)}
e^{(r-q_j)t_a+(r-q_\ell)t_b+\rho_{j\ell}\sigma_j\sigma_\ell\min(t_a,t_b)}.
\]
Then
\[
\E[B_{t_a}B_{t_b}]
=
\sum_{j=1}^d\sum_{\ell=1}^d
w_jw_\ell
\E[S_{t_a}^{(j)}S_{t_b}^{(\ell)}].
\]
Finally,
\[
\Var(\bar B)=\E[\bar B^2]-\E[\bar B]^2,
\]
where \(\E[\bar B^2]\) is obtained by summing \(\E[B_{t_a}B_{t_b}]\) over all future fixing pairs and adding deterministic running-sum terms.

\section{Moment-matched lognormal baseline}
\begin{proofsteps}
\item Suppose a positive average has mean \(m\) and variance \(v\).
\item Approximate it by \(L=e^Y\), \(Y\sim N(\mu_L,s_L^2)\).
\item The lognormal mean and variance are
\[
\E[L]=e^{\mu_L+s_L^2/2},\qquad
\Var(L)=e^{2\mu_L+s_L^2}(e^{s_L^2}-1).
\]
\item Set \(e^{\mu_L+s_L^2/2}=m\).
\item Divide the variance equation by \(m^2\):
\[
\frac{v}{m^2}=e^{s_L^2}-1.
\]
\item Hence
\[
s_L^2=\log(1+v/m^2),\qquad
\mu_L=\log m-\frac12s_L^2.
\]
\item Use the Black-Scholes-style truncated-lognormal formula to value \((L-K)_+\).
\end{proofsteps}

\section{PCA optimality}
\begin{theorem}[Rayleigh-Ritz/PCA optimality]
For a centered random vector with covariance matrix \(\Sigma\), the rank-\(k\) orthogonal projection minimizing mean-square reconstruction error projects onto the eigenvectors of \(\Sigma\) with the \(k\) largest eigenvalues.
\end{theorem}
\begin{proofsteps}
\item Let \(R\) be a centered random vector with covariance \(\Sigma\).
\item A rank-\(k\) orthogonal projection \(P\) has reconstruction error
\[
\E[\|R-PR\|^2]=\operatorname{tr}(\Sigma)-\operatorname{tr}(P\Sigma).
\]
\item Minimizing error is therefore equivalent to maximizing \(\operatorname{tr}(P\Sigma)\).
\item The Rayleigh-Ritz theorem says this trace is maximized by projecting onto the eigenvectors with the \(k\) largest eigenvalues.
\item Therefore PCA gives the best \(k\)-dimensional linear summary in mean-square reconstruction error.
\end{proofsteps}

\chapter{Line-by-Line Proofs: Sobol, Design, and Validation}

\section{Why powers of two are used}
Sobol sequences are digital nets in base two. The first \(2^m\) points have the balance property for elementary binary boxes. If we stop at an arbitrary \(N\), the last partial block may break this balance. The implementation therefore rounds a target path count to
\[
  N_{\rm Sobol}=2^{\lceil\log_2 N_{\rm target}\rceil}.
\]
This is not an unbiasedness theorem. It is a numerical integration design rule.

\section{Common random numbers}
\begin{proofsteps}
\item At state \(x_i\), the label is \(V(x_i)=\E[h(x_i,Z)]\).
\item Using the same \(Z_n\) for several states gives
\[
\widehat V(x_i)=\frac1N\sum_{n=1}^N h(x_i,Z_n).
\]
\item For each fixed \(i\), the marginal distribution of \(Z_n\) is unchanged.
\item Therefore each \(\widehat V(x_i)\) remains unbiased.
\item The errors across nearby states become correlated, which often makes label noise smoother and easier for interpolation to absorb.
\end{proofsteps}

\section{Benchmark independence}
If the same paths are used for training labels and benchmark labels, the measured proxy error can be artificially small because both contain the same random noise. Independent benchmark streams estimate
\[
  \widehat V_{\rm proxy}(x)-V_{\rm model}(x)
\]
with less leakage. When a closed-form, finite-difference, or tree benchmark exists, it is preferred because it removes benchmark Monte Carlo noise.

\chapter{Line-by-Line Proofs: SLV and Basket Cliquet}

\section{Variance process discretization}
The SLV cliquet code uses an Euler-type variance update with positivity protection. A schematic Heston-style variance step is
\[
  v_{t+\Delta t}
  =
  \max\!\left(
  v_t+\kappa(\theta-v_t)\Delta t+\xi\sqrt{v_t^+}\sqrt{\Delta t}\,Z_v,\;0
  \right).
\]
The max is not an exact theorem; it is a numerical safeguard. The consequence is that benchmark quality must be checked by increasing time steps or comparing against a stronger scheme when precision matters.

\section{Local volatility multiplier}
A local volatility multiplier \(L(S)\) changes the spot diffusion to
\[
  \frac{\dd S_t}{S_t}
  =
  (r-q)\dd t+L(S_t)\sqrt{v_t}\,\dd W_t.
\]
The pricing theorem still applies once the simulator defines the risk-neutral dynamics. The proxy must include enough state information, at least \(S_t\) and \(v_t\), because the future law depends on both.

\section{Grouped accrued labels}
For cliquets, suppose a simulated future path gives a remaining coupon sum \(C^{(n)}\). For accrued values \(a_1,\ldots,a_L\),
\[
  H_\ell^{(n)}
  =
  N_0\clip(a_\ell+C^{(n)},G_L,G_U).
\]
One simulation of \(C^{(n)}\) prices all \(L\) accrued layers. This is exact path reuse, not approximation, because the same future path sum is valid for every starting accrued value.

\section{Why basket order statistics are hard}
For a worst-of coupon,
\[
  C_k=\clip(\min_j R_k^{(j)},\ell,u).
\]
The identity of the minimum can change abruptly as the state changes. This creates nonsmooth interaction surfaces in spot and variance. A low-degree global polynomial can miss these switching surfaces. That is why the methodology allows a slower cached Sobol/LR safety proxy for strict max-error cases.

\backmatter
\chapter{References}
\begin{itemize}
  \item Black, F. and Scholes, M. (1973). The Pricing of Options and Corporate Liabilities. \emph{Journal of Political Economy}.
  \item Fritsch, F. N. and Carlson, R. E. (1980). Monotone Piecewise Cubic Interpolation. \emph{SIAM Journal on Numerical Analysis}.
  \item Fritsch, F. N. and Butland, J. (1984). A Method for Constructing Local Monotone Piecewise Cubic Interpolants. \emph{SIAM Journal on Scientific and Statistical Computing}.
  \item Hoerl, A. E. and Kennard, R. W. (1970). Ridge Regression: Biased Estimation for Nonorthogonal Problems. \emph{Technometrics}.
  \item Glasserman, P. (2004). \emph{Monte Carlo Methods in Financial Engineering}. Springer.
  \item Glasserman, P. and Staum, J. (2001). Conditioning on One-Step Survival for Barrier Option Simulations. \emph{Operations Research}.
  \item Broadie, M., Glasserman, P., and Kou, S. G. (1997). A Continuity Correction for Discrete Barrier Options. \emph{Mathematical Finance}.
\end{itemize}

\end{document}
